### 2019-05-31 Output/ModelingUpdates.Rmd
Begin log file for results summaries

---
title: "Demography Model Updates"
author: "Nick Gotelli"
date: ""
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This markdown file gives updates of progress on the demographic modeling. The most recent postings are at the top.

# 31 May 2019

![Caption text can go here.](../Graphics/Figure1.jpeg)

![Caption text here. ](../Graphics/Figure2.jpeg)

![Caption text.](../Graphics/Figure3.jpeg)

### 2019-06-02 Output/ModelingUpdates.Rmd
Add text and graphs for medfly model

@@
 knitr::opts_chunk$set(echo = TRUE)
+library(knitr)
+library(kableExtra)
+library(xtable)
+options(knitr.table.format = "latex")
 ```
@@
 
-# 31 May 2019
+# 1 June 2019
 
-![Caption text can go here.](../Graphics/Figure1.jpeg)
+## Medfly life table construction
+
+We began with the life history data  in Tables 3 and 4 of:
+
+>Vargas, R.I., W.A. Walsh, D. Kanehisa,  J.D. Stark, and T. Nishida. 2000. Annals of the Entomologial Society of America 93: 75-81.
+
+These authors raised flies at 4 different temperatures and constructed age-structured life tables with $l_x$ and $m_x$ data for deaths and births.
+Unfortunately, the full life tables were not presented, so we were forced to work with summary statistics, such as median $l_x$ values.
+We were able to assemble them into a 4-stage life history model.
+For example, here is our derived stage-transition matrix for flies raised at
+$18.5^\circ$ C:
+
+```{r, echo=FALSE}
+life_table <- matrix(c(0,0.970,0,0,0,0,0.887,0,0,0,0,1.000,1.350,0,0,0),nrow=4)
+row.names(life_table) <- c("Egg","Larva","Pupa","Adult")
+colnames(life_table) <- c("Egg","Larva","Pupa","Adult")
+kable(life_table, "latex",  booktabs = T,digits=3) %>% kable_styling(latex_options = c("striped", "hold_position"))
+```
+
+The problem with this life table is that we have forced the medfly life history into an age-structured model in which each of the 4 life history stages lasts exactly one day.
+At the end of the day, an individual either dies or advances to the next stage.
+
+Corey is working on an elegant solution to this problem that uses information on the average length of each stage.
+But for now, this is the only life table that we could construct for the medfly summary data.
+
+In our model, we have 4 such matrices (one for each temperature treatment).
+As described in the google-docs, we next pulled out the four non-zero life history parameters for this model.
+For each parameter, we fit an OLS quadratic regression model to the relationship between the life history parameter and the temperature.
+This model allows us to predict each of the 4 life history transitions as a function of temperature.
+Those 4 predictions are dropped into the matrix model, and the population growth rate $\lambda$ is derived.
+I am calling this the `vital rates model`.
+
+Because of the obvious distortions to the life history in the `vital rates model`, I decided to construct a much simpler forecasting model for these data.
+Although Vargas et al.
+(2000) do not give all the life history data we need, they do give the 4 final lambda values (one for each temperature) derived from their life history calculations.
+I used those 4 estimates in a much simpler `lambda model`.
+In this case, I simply fit the quadratic regression directly to their reported lambda values without creating or analyzing a life table.
+
+## Comparison of predictions from `vital rates model` and `lambda model`
+
+Here is a graph of the predictions from each of the two models as a function of temperature.
+The graph looks a little odd because I clipped values less than 1.0 to 0.0 so that Matt's spatial maps would be easier to read.
+I also superimposed the 4 $\lambda$ values that were used to fit the `lambda model`.
+
+![Predictions of vital rates and lambda models](../Graphics/Figure1.jpeg)
+
+There are some differences between these models.
+The `vital rates model` predicts much faster population growth rates at most temperatures, but it also generates a range of viable temperatures ($\lambda > 1.0$) that is $3^\circ$ narrower than the range predicted by the `lambda model`.
+
+In spite of this difference, the predictions from the monthly and total maps (see below) did not appear that different for the two models.
+But for our paper, I think we should stick with the `lambda model`.
+It is the very simplest demographic prediction model that is possible: no life history details are needed, just values of $\lambda$ measured at different temperatures (or other environmental variables).
+
+That should make a nice lead-in to the cheatgrass analyses, where I will again set up the `vital rates model` and the `lambda model`, this time with snowpack as the driver variable.
 

### 2019-06-02 Scripts/ExtrapolationGraph.R

@@ -45,6 +45,7 @@ model <- rep(c("Vital Rates Prediction","Lambda Prediction"),each=length(con_tem
 df <- data.frame(lambda=lambda.vec,temperature=temperature, model=model)
 df2 <- data.frame(lambda=emp_lam,temperature=exp_tmp)
 # create extrapolation plot
+xlab = expression("Temperature " ( degree*C))
 plot_1 <- ggplot(data=df, aes(x=temperature,
 															y=lambda,
 															shape=model)) +
@@ -52,12 +53,13 @@ plot_1 <- ggplot(data=df, aes(x=temperature,
 	geom_line() +
 	geom_hline(yintercept=1.0,color=I("red")) +
 	 geom_point(data=df2,color="blue",shape=21,size=4) +
+	 labs(x=xlab) +
 	theme_bw()
 
 plot(plot_1)
-ggsave(filename="Graphics/Figure1.eps",
+ggsave(filename="Graphics/Figure1.jpeg",
 			 plot=plot_1,
-			 device="eps",
+			 device="jpg",
 			 width=7,
 			 height=5,
 			 units="in")

### 2019-06-02 Output/ModelingUpdates.Rmd
Finish medfly summary

@@
 author: "Nick Gotelli"
-date: ""
-output: pdf_document
+date: ''
+output:
+  html_document:
+    theme: cerulean
+fontsize: 12pt
 ---
@@
 library(xtable)
-options(knitr.table.format = "latex")
+options(knitr.table.format = "html")
 ```
-
 This markdown file gives updates of progress on the demographic modeling. The most recent postings are at the top.
 
-# 1 June 2019
+# `1 June 2019`
 
@@
 These authors raised flies at 4 different temperatures and constructed age-structured life tables with $l_x$ and $m_x$ data for deaths and births.
-Unfortunately, the full life tables were not presented, so we were forced to work with summary statistics, such as median $l_x$ values.
-We were able to assemble them into a 4-stage life history model.
-For example, here is our derived stage-transition matrix for flies raised at
+Unfortunately, the full life tables were not presented, so I was forced to work with summary statistics, such as median $l_x$ values.
+I was able to assemble them into a 4-stage life history model.
+For example, here is the derived stage-transition matrix for flies raised at
 $18.5^\circ$ C:
@@
 colnames(life_table) <- c("Egg","Larva","Pupa","Adult")
-kable(life_table, "latex",  booktabs = T,digits=3) %>% kable_styling(latex_options = c("striped", "hold_position"))
+kable(life_table, "html",  booktabs = T,digits=3) %>% kable_styling(latex_options = c("striped", "hold_position"))
 ```
@@
 The graph looks a little odd because I clipped values less than 1.0 to 0.0 so that Matt's spatial maps would be easier to read.
-I also superimposed the 4 $\lambda$ values that were used to fit the `lambda model`.
+I also superimposed the 4 empirical $\lambda$ values (blue circles) from Vargas et al.
+(2000) that were used to fit the `lambda model`.
 
@@
 
-In spite of this difference, the predictions from the monthly and total maps (see below) did not appear that different for the two models.
+In spite of this difference, the predictions depicted in the monthly and yearly maps (see below) did not appear qualitatively different for the two models.
 But for our paper, I think we should stick with the `lambda model`.
@@
 
-![Caption text here. ](../Graphics/Figure2.jpeg)
+## SDM monthly projections with demographic model
 
-![Caption text.](../Graphics/Figure3.jpeg)
+Now that we have functions for predicting $\lambda$ as a function of temperature, we can create a simple species distribution model for medfly in North America.
+Using the bioclim variable of average monthly temperature, Matt first coded up a mapping of population growth rate in each month of the year.
+In this map, colored pixels have $\lambda > 1$, indicating sites where medfly could achieve population growth in a particular month.
+Sites in which $\lambda < 1$ have no color because these are sites where population growth rates are below replacement.
+
+Here is Matt's map:
+
+![Monthly population growth rates](../Graphics/Figure2.jpeg)
+
+From this map, you can see that the model predicts that medfly could thrive throughout most of the continent in the warm summer months, but will not persist in the winter months.
+
+Following a suggestion from Michael, we used the geometric mean of the monthly lambda values to determine those places in North America where medfly could maintain positive population growth through the course of a entire year:
+
+![Integrated annual population growth rates](../Graphics/Figure3.jpeg)
+
+Sustained population growth is only possible in south Florida; similar results were obtained using the `vital rates` model.
+From conversations with Michael and Leigh, this seems to match pretty well what has happened with medfly, which is established in south Florda, with populations in southern California that are in watered agricultural sites.
+Not bad for a demographic SDM that is based on only 4 data points from a lab study!
+
+If there are GBIF records available for medfly, it would be interesting to compare these results to a cookie-cutter MaxEnt model that uses the default settings (Matt, would this be easy to do?).
+
+The next step will be to start setting up the cheatgrass models.

### 2019-06-03 Scripts/MatrixFunctions.R
Correct clipping in graphics

@@ -96,8 +96,8 @@ mat_fill <- function(z=1) { # place predicted coefficients in new transition mat
 #-------------------------------------
 t2 <- map(t1,mat_fill) %>% # extract population growth rate from 
 map(lambda) %>%
-# t3 <- lapply(t2, function (x) ifelse(x < 1,0,x))
-map(~ifelse(. < 1,0,.))
+# t3 <- lapply(t2, function (x) ifelse(x < 0,0,x))
+map(~ifelse(. < 0,0,.))# rescale any values with lambda <0 to 0
 return(t2) 
 
 }
@@ -110,7 +110,7 @@ return(t2)
 lam2_gen <- function(x,y,z){
 . <- est_trans(x,y)
 . <- pred_trans(z,.)
-.[.<1] <- 0  # rescale any values with lambda <1 to 0
+.[.<0] <- 0  # rescale any values with lambda <0 to 0
  names(.) <- paste0("V",seq_along(z))
 . <- as.list(.)
 return(.)

### 2019-06-03 Scripts/MattsGraphicsCode.R

@@ -66,11 +66,13 @@ lambda_rasts <- mclapply(1:12, function(x, tempRasts){
   rast <- tempRasts[[x]]
   keep <- !is.na(rast[]) # index NAs
 #-------------------------------------
-   # lamb <- lam_gen(rast[keep]) # calc lamba for each temp
+  lamb <- lam_gen(rast[keep]) # calc lamba for each temp
+  lamb[lamb<1] <-0 # clip non-increasing values to 0
 #-------------------------------------
-  lamb <- lam2_gen(x=exp_tmp,
-  								 y=emp_lam,
-  								 z=rast[keep]) # calc lamba for each temp
+  # lamb <- lam2_gen(x=exp_tmp,
+  # 								 y=emp_lam,
+  # 								 z=rast[keep]) # calc lamba for each temp
+  # lamb[lamb<1] <-0 # clip non-increasing values to 0
 #-------------------------------------
   rast[which(keep)] <- unlist(lamb) # assign lamba back to raster
   return(rast)}, tempRasts=tmean, mc.cores=2)

### 2019-07-18 Scripts/BromusLambdaModel.R
Add clean Bromus data set and begin quadratic model fitting

@@ -1,0 +1,30 @@
+# Analyze cheatgrass data for regression model
+# of lambda versus snow, precip, and temperature
+# Original data from Alden Griffith and Michael Loic
+#
+# NJG
+# 17 July 2019
+
+# Preliminaries -----------------------------------------------------------
+library(tidyverse)
+
+# Get input data ----------------------------------------------------------
+
+df <- read.csv(file="CleanData/CleanBromusData.csv",comment.char="#")
+df <- df[,-ncol(df)]
+df_clean <- df[complete.cases(df),]
+qplot(df_clean$Lambda, color=I("black"), fill=I("tan"))
+qplot(y=df_clean$Lambda,x=df_clean$Snow.Cover)
+
+
+# Function Model Fit ------------------------------------------------------
+# inputs: response variable, predictor variable
+# outputs: adjusted r2 for linear and quadratic regression models
+
+quad_fit <- function(x=df_clean$Snow.Cover, y=df_clean$Lambda){
+  M1 <- summary(lm(y~x))
+  M2 <- summary(lm(y~poly(x,2)))
+  result <- list(M1,M2)
+}
+z <- quad_fit()
+print(z)

### 2019-07-20 Scripts/BromusUpdate.Rmd
Create progress report for Bromus analyses

---
title: "*Bromus* Initial Analysis"
author: "Nick Gotelli"
date: "7/19/2019"
output: 
  html_document: 
    highlight: tango
    theme: cerulean
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data sources and cleaning
These data were provided by Alden Griffith and Michael Loic. After talking with Alden, we decided to use lambda values that had been estimated from changes in the number of individuals in marked plots, rather than analysis of stage transition matrices. For a number of control and snow-manipulated plots, we have the estimated lambda $(N_{t + 1}/N_{t})$ based on annual counts of cheatgrass for 7 consecutive years.

I cleaned the initial data file (`Bromus Lambda and Climate Data.csv`), moving a notes column to initial comment lines, and adding 4 missing values as `NA`. The modified data are in `CleanBromusData.csv`.

```{r}
# Preliminaries -----------------------------------------------------------
library(tidyverse)

# Get input data ----------------------------------------------------------
df <- read.csv(file="../CleanData/CleanBromusData.csv",comment.char="#")
df <- df[,-ncol(df)]                 # remove empty final column
df_clean <- df[complete.cases(df),]  # remove 4 missing lambda

df_preds <- df_clean[,8:18]           # create dataframe of predictor variables
```

## Initial regression models
With the data frame set up, the first analysis was to use each of the predictor variables (snow, degree-days, precipitation, etc.), with $\lambda$ as the response variable. For each variable, I built a linear and a quadratic regression model.

```{r}
# Function Model Fit ------------------------------------------------------
# inputs: response variable, predictor variable
# outputs: adjusted r2 for linear and quadratic regression models

quad_fit <- function(x=df_clean$Snow.Cover, y=df_clean$Lambda){
  M1 <- summary(lm(y~x))
  M2 <- summary(lm(y~poly(x,2)))
  return(c(M1$adj.r.squared,M2$adj.r.squared))
}
# -------------------------------------------------------------------------
# map all predictor variables to quad_fit
summary_df <- map(df_preds,quad_fit,y=df_clean$Lambda)

# put into a summary data frame and add column labels
summary_df <- t(bind_rows(summary_df))
colnames(summary_df) <- c("Linear","Quadratic")
# -------------------------------------------------------------------------
```

## First screening of predictor variables
In this table, each row is a different model with a single predictor variable, and the data frame shows the adjusted r^2^ for each model. The two columns are for a linear model $(y \sim x)$ and a quadratic model $(y \sim x + x^2)$.

```{r}
print(summary_df)
```

A couple things jump out from these univariate regressions. First, the quadratic model almost always fits better than the corresponding linear model (even after imposing a penalty for the additional parameter), so we have good theoretical grounds and empirical support for going with the quadratic models.

I wanted to have predictors related to snow, temperature, and precipitation, as those seem like the best suspects for what is driving distribution. Based on the adjusted r^2^, the best 3 predictors were `Snow.Cover`, `Grow.Season.Precip.Total`, and `Grow.Season.DD.all.days`. There was also a significant effect of `Starting.Density`, which I will discuss near the end.

## Initial multiple regression model

Now taking those 3 best predictors based on snow, precip, and temperature, here is the full model:

```{r}
# best variables are:
# Grow.Season.DD.all.days (4)
# Grow.Season.Precip.Total (5)
# Snow.Cover (6)
best_preds <- df_preds[,4:6]

y <- df_clean$Lambda
x_1 <- best_preds[,1]
x_2 <- best_preds[,2]
x_3 <- best_preds[,3]

model_1 <- summary(lm(y~poly(x_1,2) + poly(x_2,2) + poly(x_3,2)))
print(model_1)

```

## 2-variable model with an interaction term

Although the precipitation variable was initially a strong predictor, its effect is non-significant with both snow cover and degree days in the model, so I dropped it at this point. The next step was to try a 2-variable model but include an interaction term between snow cover and degree days:

```{r}
# Try adding an interaction term -----------------------------------------

model_3 <- summary(lm(y~poly(x_1,2) + poly(x_3,2) + x_1*x_3))
print(model_3)
```

## Final 2-variable prediction model

However, the interaction term does not add anything. So, we are left with just a two-variable quadratic model for predicting $\lambda$:

```{r}
model_2 <- summary(lm(y~poly(x_1,2) + poly(x_3,2)))
print(model_2)

```
In this model, all of the coefficients are significantly different from zero, and the model explains over 50\% of the variance in $\lambda$, which is decent. For now, that is the model we will start with, although it will be easy enough to change if there are other variables you want to consider.

## Functions for fitting and predicting with new values

The next two functions are used to fit the model to the data, and then to predict new values of $\lambda$ based on new inputs of snow cover or degree days.

```{r}
# Function Lambda Model -----------------------------------------------
# inputs: lambda vector, and data frame of predictor variables
# outputs: linear model for forecasting

lambda_model <- function(y=df_clean$Lambda,z=best_preds){
  y <- df_clean$Lambda
  x_1 <- z[,1] # Grow.Season.DD.all.days
  x_2 <- z[,3]  # Snow.Cover
  forecast_model <- lm(y~poly(x_1,2) + poly(x_2,2))
}

# Function Lambda predict -------------------------------------------------
# inputs: linear model, degree days (vector), snow cover (vector)
# outputs: predicted lambda

lambda_predict <- function(m=my_model,degree_days=c(800,600),snow=c(0.83,.74)) {

z <- predict(object=m, newdata=data.frame(x_1=degree_days,x_2=snow))
z[z<0] <- 0   # reset to 0 any predicted negative values for lambda
return(z)
}

```

## Using the functions for forecasting continental distributions

These two functions will make it possible to generate predictions for any values of snow cover and degree days. For example, here the functions are run for two arbitrary vectors of 10 elements each snow cover and degree days. The output is the estimated $\lambda$ for each pair of the variables:

```{r}
lambda_predict(m=lambda_model(),degree_days=sample(600:800,10),snow=runif(10))
```

When we get to the next step, Matt will use this line of code to generate an estimated lambda for any map location for which we have the snow and degree days measurements.

## Sweeping the parameters and constructing a heat map for $\lambda$

Before we build the continental map, it is instructive to sweep the parameters and visualize the response surface of $\lambda$. To set limits on snow cover and degree days, I will consider snow cover between 0 and 100%, and will expand the degree day measurements by 50% beyond the measured limits in Alden and Michael's experiment:

```{r}
# set limits for parameter inputs
snow_low <- 0.0
snow_high <- 1.0
dd_low <- 0.5*min(x_1)
dd_high <- 1.5*max(x_1)

# create parameter data frame with column for output
param_df <- expand.grid(snow_cover=seq(from=snow_low, to=snow_high, length.out=20),
                        degree_days=seq(from=dd_low, to=dd_high,length.out=20))
param_df$lambda <- NA

# apply function to each row
for (i in 1:nrow(param_df)){
  param_df[i,3] <- lambda_predict(m=lambda_model(),
                                  degree_days=param_df[i,2],
                                  snow=param_df[i,1])
}

```

## Problems with the model
Scanning through the completed `param_df` data frame reveals impossibly high values of lambda for most parameter combinations. The problem is coming from the degree days data:

```{r}
qplot(x=best_preds[,1],y=df_clean$Lambda)

```

Because of the cluster of small values at 800 degree days, the quadratic term in the regression model has a positive coefficient, so $\lambda$ increases at both high and low degree days.

I don't see an easy solution, so for now, I am going to drop back to a model just using snow cover, which has a negative coefficient for the quadratic term and therefore decreases lambda at more extreme values.

## Simple snow cover model
So, here are the reworked functions just using snow cover:
```{r}
# Function Lambda Model2-----------------------------------------------
# inputs: lambda vector, and data frame of predictor variables
# outputs: linear model for forecasting

lambda_model2 <- function(y=df_clean$Lambda,z=best_preds){
  y <- df_clean$Lambda
  # x_1 <- z[,1] # Grow.Season.DD.all.days
  x_2 <- z[,3]  # Snow.Cover
  forecast_model <- lm(y~ poly(x_2,2))
}

# Function Lambda predict2 -------------------------------------------------
# inputs: linear model, degree days (vector), snow cover (vector)
# outputs: predicted lambda

lambda_predict2 <- function(m=my_model,snow=c(0.83,.74)) {

z <- predict(object=m, newdata=data.frame(x_2=snow))
z[z<0] <- 0   # reset to 0 any predicted negative values for lambda
return(z)
}

param_df2 <- data.frame(snow_cover=seq(from=snow_low, to=snow_high, length.out=20),
                        lambda=NA)
```

Now with these simplified functions, we run the model once again:

```{r}

# apply function to each row
for (i in 1:nrow(param_df2)){
  param_df2[i,2] <- lambda_predict2(m=lambda_model2(),
                                  snow=param_df2[i,1])
}
plot_1 <- ggplot(param_df2,aes(x=snow_cover,y=lambda)) +
          geom_point() +
          xlab("Snow Cover") +
          ylab("Predicted Lambda")

plot(plot_1)
```

These results look much more reasonable. The maximum predicted $\lambda$ is only 2.5, which is still high, but not unreasonable. And positive population growth ($\lambda > 1$) is predicted for snow cover values from \~ 35% to 99%.

And, for comparison, here is a plot of the original data from which the prediction equation was derived:

```{r}
df <- data.frame(Snow.Cover=df_clean$Snow.Cover, Lambda=df_clean$Lambda)
plot_2 <- ggplot(df,aes(x=Snow.Cover,y=Lambda)) +
          geom_point() +
          xlab("Snow Cover") +
          ylab("Observed Lambda")

plot(plot_2)
```

## Remaining questions and issues

1. Obviously, the biggest problem is what to do about these environmental variables like degree days that do not have well-behaved quadratic functions. I wonder if this is an artifact caused by snow cover being manipulated, but environmental variables not being controlled. However, I thought that Alden had provided me with just data from all the control plots, so I don't know if this is the issue. I was really hoping we could have at least two predictor variables in the regression model, but that may not work for this data set.

2. In order for Matt to use this model for continental projections, we will need snow cover for each pixel in the continental landscape. From talking with Alden, it sounds like those data will be available, although I don't know much about the grain sizes and resolution.

3. In the first set of regression models, there is a strong effect of initial density, suggesting density dependence, but both the linear and quadratic coefficients are positive, so this is actually some form of inverse density dependence. I can't tell if this is an artifact, or if the density effects are what is actually driving the other patterns that we are seeing. But we need to discuss this more.

4. Alden's data set has additional columns indicating snow conditions (Upwind, downwind, plus, minus). I have ignored those designations in this analysis, but they obviously contribute to variation in $\lambda$. If you want, it would be easy enough to go back and subscript the data to fit to fewer points. That will reduce the variance, although I don't think it is going to change the shape of the curves that much.

5. As you can see, the basic model with snow cover looks reasonable, but we need to figure out what, if anything to do with the other environmental variables. Once we get this sorted out and agree on a final regression model, I will clean up the code so that Matt has a single function to work with for generating maps.

6. Michael L. and Alden (and Jen??), we should probably talk again on skype after you have digested these results. I will hold off on any other analyses until we have talked.

### 2019-07-20 Scripts/BromusLambdaModel.R

@@ -9,14 +9,11 @@
 library(tidyverse)
 
 # Get input data ----------------------------------------------------------
-
 df <- read.csv(file="CleanData/CleanBromusData.csv",comment.char="#")
-df <- df[,-ncol(df)]
-df_clean <- df[complete.cases(df),]
-qplot(df_clean$Lambda, color=I("black"), fill=I("tan"))
-qplot(y=df_clean$Lambda,x=df_clean$Snow.Cover)
+df <- df[,-ncol(df)]                 # remove empty final column
+df_clean <- df[complete.cases(df),]  # remove 4 missing lambda
 
-
+df_preds <- df_clean[,8:18]           # create dataframe of predictor variables
 # Function Model Fit ------------------------------------------------------
 # inputs: response variable, predictor variable
 # outputs: adjusted r2 for linear and quadratic regression models
@@ -24,7 +21,72 @@ qplot(y=df_clean$Lambda,x=df_clean$Snow.Cover)
 quad_fit <- function(x=df_clean$Snow.Cover, y=df_clean$Lambda){
   M1 <- summary(lm(y~x))
   M2 <- summary(lm(y~poly(x,2)))
-  result <- list(M1,M2)
+  return(c(M1$adj.r.squared,M2$adj.r.squared))
 }
-z <- quad_fit()
-print(z)
+# -------------------------------------------------------------------------
+# map all predictor variables to quad_fit
+summary_df <- map(df_preds,quad_fit,y=df_clean$Lambda)
+
+# put into a summary data frame and add column labels
+summary_df <- t(bind_rows(summary_df))
+colnames(summary_df) <- c("Linear","Quadratic")
+# -------------------------------------------------------------------------
+
+print(summary_df)
+
+
+# Initial multiple regression model fit ----------------------------------
+
+# best variables are:
+# Grow.Season.DD.all.days (4)
+# Grow.Season.Precip.Total (5)
+# Snow.Cover (6)
+best_preds <- df_preds[,4:6]
+  
+y <- df_clean$Lambda
+x_1 <- best_preds[,1]
+x_2 <- best_preds[,2]
+x_3 <- best_preds[,3]
+
+model_1 <- summary(lm(y~poly(x_1,2) + poly(x_2,2) + poly(x_3,2)))
+print(model_1)
+
+
+# Simplified best fit model -----------------------------------------------
+
+model_2 <- summary(lm(y~poly(x_1,2) + poly(x_3,2)))
[…]

### 2019-07-23 Scripts/BromusUpdate.Rmd
Checking for LaTeX installation

@@
 ---
-title: "*Bromus* Initial Analysis"
+title: '*Bromus* Initial Analysis'
 author: "Nick Gotelli"
@@
 output:
+  pdf_document: default
   html_document:

### 2019-07-23 Scripts/BromusUpdate.tex
Checking for LaTeX installation

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Bromus Initial Analysis},
            pdfauthor={Nick Gotelli},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\providecommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}

  \title{\emph{Bromus} Initial Analysis}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{Nick Gotelli}
    \preauthor{\centering\large\emph}
  \postauthor{\par}
      \predate{\centering\large\emph}
  \postdate{\par}
    \date{7/19/2019}

\begin{document}
\maketitle

\hypertarget{data-sources-and-cleaning}{%
\subsection{Data sources and cleaning}\label{data-sources-and-cleaning}}

These data were provided by Alden Griffith and Michael Loic. After
talking with Alden, we decided to use lambda values that had been
estimated from changes in the number of individuals in marked plots,
rather than analysis of stage transition matrices. For a number of
control and snow-manipulated plots, we have the estimated lambda
\((N_{t + 1}/N_{t})\) based on annual counts of cheatgrass for 7
consecutive years.

I cleaned the initial data file
(\texttt{Bromus\ Lambda\ and\ Climate\ Data.csv}), moving a notes column
to initial comment lines, and adding 4 missing values as \texttt{NA}.
The modified data are in \texttt{CleanBromusData.csv}.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# Preliminaries -----------------------------------------------------------}
\KeywordTok{library}\NormalTok{(tidyverse)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## -- Attaching packages -------------------- tidyverse 1.2.1 --
\end{verbatim}

\begin{verbatim}
## v ggplot2 3.2.0     v purrr   0.3.2
## v tibble  2.1.3     v dplyr   0.8.3
## v tidyr   0.8.3     v stringr 1.4.0
## v readr   1.3.1     v forcats 0.4.0
\end{verbatim}

\begin{verbatim}
## -- Conflicts ----------------------- tidyverse_conflicts() --
## x dplyr::filter() masks stats::filter()
## x dplyr::lag()    masks stats::lag()
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# Get input data ----------------------------------------------------------}
\NormalTok{df <-}\StringTok{ }\KeywordTok{read.csv}\NormalTok{(}\DataTypeTok{file=}\StringTok{"../CleanData/CleanBromusData.csv"}\NormalTok{,}\DataTypeTok{comment.char=}\StringTok{"#"}\NormalTok{)}
\NormalTok{df <-}\StringTok{ }\NormalTok{df[,}\OperatorTok{-}\KeywordTok{ncol}\NormalTok{(df)]                 }\CommentTok{# remove empty final column}
\NormalTok{df_clean <-}\StringTok{ }\NormalTok{df[}\KeywordTok{complete.cases}\NormalTok{(df),]  }\CommentTok{# remove 4 missing lambda}

\NormalTok{df_preds <-}\StringTok{ }\NormalTok{df_clean[,}\DecValTok{8}\OperatorTok{:}\DecValTok{18}\NormalTok{]           }\CommentTok{# create dataframe of predictor variables}
\end{Highlighting}
\end{Shaded}

\hypertarget{initial-regression-models}{%
\subsection{Initial regression models}\label{initial-regression-models}}

With the data frame set up, the first analysis was to use each of the
predictor variables (snow, degree-days, precipitation, etc.), with
\(\lambda\) as the response variable. For each variable, I built a
linear and a quadratic regression model.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# Function Model Fit ------------------------------------------------------}
\CommentTok{# inputs: response variable, predictor variable}
\CommentTok{# outputs: adjusted r2 for linear and quadratic regression models}

\NormalTok{quad_fit <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(}\DataTypeTok{x=}\NormalTok{df_clean}\OperatorTok{$}\NormalTok{Snow.Cover, }\DataTypeTok{y=}\NormalTok{df_clean}\OperatorTok{$}\NormalTok{Lambda)\{}
\NormalTok{  M1 <-}\StringTok{ }\KeywordTok{summary}\NormalTok{(}\KeywordTok{lm}\NormalTok{(y}\OperatorTok{~}\NormalTok{x))}
\NormalTok{  M2 <-}\StringTok{ }\KeywordTok{summary}\NormalTok{(}\KeywordTok{lm}\NormalTok{(y}\OperatorTok{~}\KeywordTok{poly}\NormalTok{(x,}\DecValTok{2}\NormalTok{)))}
  \KeywordTok{return}\NormalTok{(}\KeywordTok{c}\NormalTok{(M1}\OperatorTok{$}\NormalTok{adj.r.squared,M2}\OperatorTok{$}\NormalTok{adj.r.squared))}
\NormalTok{\}}
\CommentTok{# -------------------------------------------------------------------------}
\CommentTok{# map all predictor variables to quad_fit}
\NormalTok{summary_df <-}\StringTok{ }\KeywordTok{map}\NormalTok{(df_preds,quad_fit,}\DataTypeTok{y=}\NormalTok{df_clean}\OperatorTok{$}\NormalTok{Lambda)}

\CommentTok{# put into a summary data frame and add column labels}
\NormalTok{summary_df <-}\StringTok{ }\KeywordTok{t}\NormalTok{(}\KeywordTok{bind_rows}\NormalTok{(summary_df))}
\KeywordTok{colnames}\NormalTok{(summary_df) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"Linear"}\NormalTok{,}\StringTok{"Quadratic"}\NormalTok{)}
\CommentTok{# -------------------------------------------------------------------------}
\end{Highlighting}
\end{Shaded}

\hypertarget{first-screening-of-predictor-variables}{%
\subsection{First screening of predictor
variables}\label{first-screening-of-predictor-variables}}

In this table, each row is a different model with a single predictor
variable, and the data frame shows the adjusted r\textsuperscript{2} for
each model. The two columns are for a linear model \((y \sim x)\) and a
quadratic model \((y \sim x + x^2)\).

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{print}\NormalTok{(summary_df)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##                            Linear Quadratic
## Starting.Density          0.37338   0.42397
## Grow.Season.Precip.Snow   0.02884   0.04209
## Grow.Season.Precip.Rain   0.05091   0.39264
## Grow.Season.DD.all.days   0.20921   0.34622
## Grow.Season.Precip.Total  0.09829   0.47520
## Snow.Cover                0.22703   0.45274
## Winter.Precip.Snow        0.05769   0.15705
## Winter.Precip.Rain        0.42714   0.42819
## Winter.DD.no.snowpack     0.00414  -0.00406
## Winter.DD.all.days       -0.00690   0.14229
## Winter.Precip.Total       0.14253   0.28153
\end{verbatim}

A couple things jump out from these univariate regressions. First, the
quadratic model almost always fits better than the corresponding linear
model (even after imposing a penalty for the additional parameter), so
we have good theoretical grounds and empirical support for going with
the quadratic models.

I wanted to have predictors related to snow, temperature, and
precipitation, as those seem like the best suspects for what is driving
distribution. Based on the adjusted r\textsuperscript{2}, the best 3
predictors were \texttt{Snow.Cover}, \texttt{Grow.Season.Precip.Total},
and \texttt{Grow.Season.DD.all.days}. There was also a significant
effect of \texttt{Starting.Density}, which I will discuss near the end.

\hypertarget{initial-multiple-regression-model}{%
\subsection{Initial multiple regression
model}\label{initial-multiple-regression-model}}

Now taking those 3 best predictors based on snow, precip, and
temperature, here is the full model:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# best variables are:}
\CommentTok{# Grow.Season.DD.all.days (4)}
\CommentTok{# Grow.Season.Precip.Total (5)}
\CommentTok{# Snow.Cover (6)}
\NormalTok{best_preds <-}\StringTok{ }\NormalTok{df_preds[,}\DecValTok{4}\OperatorTok{:}\DecValTok{6}\NormalTok{]}

\NormalTok{y <-}\StringTok{ }\NormalTok{df_clean}\OperatorTok{$}\NormalTok{Lambda}
\NormalTok{x_}\DecValTok{1}\NormalTok{ <-}\StringTok{ }\NormalTok{best_preds[,}\DecValTok{1}\NormalTok{]}
\NormalTok{x_}\DecValTok{2}\NormalTok{ <-}\StringTok{ }\NormalTok{best_preds[,}\DecValTok{2}\NormalTok{]}
\NormalTok{x_}\DecValTok{3}\NormalTok{ <-}\StringTok{ }\NormalTok{best_preds[,}\DecValTok{3}\NormalTok{]}

\NormalTok{model_}\DecValTok{1}\NormalTok{ <-}\StringTok{ }\KeywordTok{summary}\NormalTok{(}\KeywordTok{lm}\NormalTok{(y}\OperatorTok{~}\KeywordTok{poly}\NormalTok{(x_}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{) }\OperatorTok{+}\StringTok{ }\KeywordTok{poly}\NormalTok{(x_}\DecValTok{2}\NormalTok{,}\DecValTok{2}\NormalTok{) }\OperatorTok{+}\StringTok{ }\KeywordTok{poly}\NormalTok{(x_}\DecValTok{3}\NormalTok{,}\DecValTok{2}\NormalTok{)))}
\KeywordTok{print}\NormalTok{(model_}\DecValTok{1}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## 
## Call:
## lm(formula = y ~ poly(x_1, 2) + poly(x_2, 2) + poly(x_3, 2))
## 
## Residuals:
##    Min     1Q Median     3Q    Max 
## -2.114 -0.454 -0.113  0.246  3.516 
## 
## Coefficients:
##               Estimate Std. Error t value Pr(>|t|)
## (Intercept)     1.6351     0.0789   20.72  < 2e-16
## poly(x_1, 2)1   3.1085     1.2259    2.54  0.01254
## poly(x_1, 2)2   4.4275     1.2023    3.68  0.00035
## poly(x_2, 2)1   0.5263     2.0579    0.26  0.79858
## poly(x_2, 2)2   2.0431     1.6022    1.28  0.20478
## poly(x_3, 2)1   5.5628     1.2775    4.35  2.9e-05
## poly(x_3, 2)2  -5.1611     1.8141   -2.84  0.00524
## 
## Residual standard error: 0.879 on 117 degrees of freedom
## Multiple R-squared:  0.612,  Adjusted R-squared:  0.592 
## F-statistic: 30.8 on 6 and 117 DF,  p-value: <2e-16
\end{verbatim}

\hypertarget{variable-model-with-an-interaction-term}{%
\subsection{2-variable model with an interaction
term}\label{variable-model-with-an-interaction-term}}

Although the precipitation variable was initially a strong predictor,
its effect is non-significant with both snow cover and degree days in
the model, so I dropped it at this point. The next step was to try a
2-variable model but include an interaction term between snow cover and
degree days:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# Try adding an interaction term -----------------------------------------}

\NormalTok{model_}\DecValTok{3}\NormalTok{ <-}\StringTok{ }\KeywordTok{summary}\NormalTok{(}\KeywordTok{lm}\NormalTok{(y}\OperatorTok{~}\KeywordTok{poly}\NormalTok{(x_}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{) }\OperatorTok{+}\StringTok{ }\KeywordTok{poly}\NormalTok{(x_}\DecValTok{3}\NormalTok{,}\DecValTok{2}\NormalTok{) }\OperatorTok{+}\StringTok{ }\NormalTok{x_}\DecValTok{1}\OperatorTok{*}\NormalTok{x_}\DecValTok{3}\NormalTok{))}
\KeywordTok{print}\NormalTok{(model_}\DecValTok{3}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## 
## Call:
## lm(formula = y ~ poly(x_1, 2) + poly(x_3, 2) + x_1 * x_3)
## 
## Residuals:
##    Min     1Q Median     3Q    Max 
## -2.113 -0.392 -0.039  0.297  3.602 
## 
## Coefficients: (2 not defined because of singularities)
##               Estimate Std. Error t value Pr(>|t|)
## (Intercept)    5.19694    3.99197    1.30  0.19550
## poly(x_1, 2)1  8.90080    5.41854    1.64  0.10312
## poly(x_1, 2)2  4.79181    0.97537    4.91  2.9e-06
## poly(x_3, 2)1 20.80155   15.83877    1.31  0.19162
## poly(x_3, 2)2 -4.23724    1.14679   -3.69  0.00033
## x_1                 NA         NA      NA       NA
## x_3                 NA         NA      NA       NA
## x_1:x_3       -0.00691    0.00775   -0.89  0.37398
## 
## Residual standard error: 0.879 on 118 degrees of freedom
## Multiple R-squared:  0.609,  Adjusted R-squared:  0.592 
## F-statistic: 36.8 on 5 and 118 DF,  p-value: <2e-16
\end{verbatim}

\hypertarget{final-2-variable-prediction-model}{%
\subsection{Final 2-variable prediction
model}\label{final-2-variable-prediction-model}}

However, the interaction term does not add anything. So, we are left
with just a two-variable quadratic model for predicting \(\lambda\):

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{model_}\DecValTok{2}\NormalTok{ <-}\StringTok{ }\KeywordTok{summary}\NormalTok{(}\KeywordTok{lm}\NormalTok{(y}\OperatorTok{~}\KeywordTok{poly}\NormalTok{(x_}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{) }\OperatorTok{+}\StringTok{ }\KeywordTok{poly}\NormalTok{(x_}\DecValTok{3}\NormalTok{,}\DecValTok{2}\NormalTok{)))}
\KeywordTok{print}\NormalTok{(model_}\DecValTok{2}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## 
## Call:
## lm(formula = y ~ poly(x_1, 2) + poly(x_3, 2))
## 
## Residuals:
##    Min     1Q Median     3Q    Max 
## -2.095 -0.428 -0.078  0.290  3.541 
## 
## Coefficients:
##               Estimate Std. Error t value Pr(>|t|)
## (Intercept)     1.6351     0.0788   20.74  < 2e-16
## poly(x_1, 2)1   4.1393     0.9448    4.38  2.6e-05
## poly(x_1, 2)2   4.8334     0.9734    4.97  2.3e-06
## poly(x_3, 2)1   6.6919     0.9478    7.06  1.2e-10
## poly(x_3, 2)2  -4.7814     0.9704   -4.93  2.7e-06
## 
## Residual standard error: 0.878 on 119 degrees of freedom
## Multiple R-squared:  0.606,  Adjusted R-squared:  0.593 
## F-statistic: 45.8 on 4 and 119 DF,  p-value: <2e-16
\end{verbatim}

In this model, all of the coefficients are significantly different from
zero, and the model explains over 50\% of the variance in \(\lambda\),
which is decent. For now, that is the model we will start with, although
it will be easy enough to change if there are other variables you want
to consider.

\hypertarget{functions-for-fitting-and-predicting-with-new-values}{%
\subsection{Functions for fitting and predicting with new
values}\label{functions-for-fitting-and-predicting-with-new-values}}

The next two functions are used to fit the model to the data, and then
to predict new values of \(\lambda\) based on new inputs of snow cover
or degree days.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# Function Lambda Model -----------------------------------------------}
\CommentTok{# inputs: lambda vector, and data frame of predictor variables}
\CommentTok{# outputs: linear model for forecasting}

\NormalTok{lambda_model <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(}\DataTypeTok{y=}\NormalTok{df_clean}\OperatorTok{$}\NormalTok{Lambda,}\DataTypeTok{z=}\NormalTok{best_preds)\{}
\NormalTok{  y <-}\StringTok{ }\NormalTok{df_clean}\OperatorTok{$}\NormalTok{Lambda}
\NormalTok{  x_}\DecValTok{1}\NormalTok{ <-}\StringTok{ }\NormalTok{z[,}\DecValTok{1}\NormalTok{] }\CommentTok{# Grow.Season.DD.all.days}
\NormalTok{  x_}\DecValTok{2}\NormalTok{ <-}\StringTok{ }\NormalTok{z[,}\DecValTok{3}\NormalTok{]  }\CommentTok{# Snow.Cover}
\NormalTok{  forecast_model <-}\StringTok{ }\KeywordTok{lm}\NormalTok{(y}\OperatorTok{~}\KeywordTok{poly}\NormalTok{(x_}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{) }\OperatorTok{+}\StringTok{ }\KeywordTok{poly}\NormalTok{(x_}\DecValTok{2}\NormalTok{,}\DecValTok{2}\NormalTok{))}
\NormalTok{\}}

\CommentTok{# Function Lambda predict -------------------------------------------------}
\CommentTok{# inputs: linear model, degree days (vector), snow cover (vector)}
\CommentTok{# outputs: predicted lambda}

\NormalTok{lambda_predict <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(}\DataTypeTok{m=}\NormalTok{my_model,}\DataTypeTok{degree_days=}\KeywordTok{c}\NormalTok{(}\DecValTok{800}\NormalTok{,}\DecValTok{600}\NormalTok{),}\DataTypeTok{snow=}\KeywordTok{c}\NormalTok{(}\FloatTok{0.83}\NormalTok{,.}\DecValTok{74}\NormalTok{)) \{}

\NormalTok{z <-}\StringTok{ }\KeywordTok{predict}\NormalTok{(}\DataTypeTok{object=}\NormalTok{m, }\DataTypeTok{newdata=}\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{x_1=}\NormalTok{degree_days,}\DataTypeTok{x_2=}\NormalTok{snow))}
\NormalTok{z[z}\OperatorTok{<}\DecValTok{0}\NormalTok{] <-}\StringTok{ }\DecValTok{0}   \CommentTok{# reset to 0 any predicted negative values for lambda}
\KeywordTok{return}\NormalTok{(z)}
\NormalTok{\}}
\end{Highlighting}
\end{Shaded}

\hypertarget{using-the-functions-for-forecasting-continental-distributions}{%
\subsection{Using the functions for forecasting continental
distributions}\label{using-the-functions-for-forecasting-continental-distributions}}

These two functions will make it possible to generate predictions for
any values of snow cover and degree days. For example, here the
functions are run for two arbitrary vectors of 10 elements each snow
cover and degree days. The output is the estimated \(\lambda\) for each
pair of the variables:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{lambda_predict}\NormalTok{(}\DataTypeTok{m=}\KeywordTok{lambda_model}\NormalTok{(),}\DataTypeTok{degree_days=}\KeywordTok{sample}\NormalTok{(}\DecValTok{600}\OperatorTok{:}\DecValTok{800}\NormalTok{,}\DecValTok{10}\NormalTok{),}\DataTypeTok{snow=}\KeywordTok{runif}\NormalTok{(}\DecValTok{10}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##     1     2     3     4     5     6     7     8     9    10 
## 1.252 2.233 2.048 0.000 2.294 4.602 1.599 0.000 0.000 0.594
\end{verbatim}

When we get to the next step, Matt will use this line of code to
generate an estimated lambda for any map location for which we have the
snow and degree days measurements.

\hypertarget{sweeping-the-parameters-and-constructing-a-heat-map-for-lambda}{%
\subsection{\texorpdfstring{Sweeping the parameters and constructing a
heat map for
\(\lambda\)}{Sweeping the parameters and constructing a heat map for \textbackslash{}lambda}}\label{sweeping-the-parameters-and-constructing-a-heat-map-for-lambda}}

Before we build the continental map, it is instructive to sweep the
parameters and visualize the response surface of \(\lambda\). To set
limits on snow cover and degree days, I will consider snow cover between
0 and 100\%, and will expand the degree day measurements by 50\% beyond
the measured limits in Alden and Michael's experiment:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# set limits for parameter inputs}
\NormalTok{snow_low <-}\StringTok{ }\FloatTok{0.0}
\NormalTok{snow_high <-}\StringTok{ }\FloatTok{1.0}
\NormalTok{dd_low <-}\StringTok{ }\FloatTok{0.5}\OperatorTok{*}\KeywordTok{min}\NormalTok{(x_}\DecValTok{1}\NormalTok{)}
\NormalTok{dd_high <-}\StringTok{ }\FloatTok{1.5}\OperatorTok{*}\KeywordTok{max}\NormalTok{(x_}\DecValTok{1}\NormalTok{)}

\CommentTok{# create parameter data frame with column for output}
\NormalTok{param_df <-}\StringTok{ }\KeywordTok{expand.grid}\NormalTok{(}\DataTypeTok{snow_cover=}\KeywordTok{seq}\NormalTok{(}\DataTypeTok{from=}\NormalTok{snow_low, }\DataTypeTok{to=}\NormalTok{snow_high, }\DataTypeTok{length.out=}\DecValTok{20}\NormalTok{),}
                        \DataTypeTok{degree_days=}\KeywordTok{seq}\NormalTok{(}\DataTypeTok{from=}\NormalTok{dd_low, }\DataTypeTok{to=}\NormalTok{dd_high,}\DataTypeTok{length.out=}\DecValTok{20}\NormalTok{))}
\NormalTok{param_df}\OperatorTok{$}\NormalTok{lambda <-}\StringTok{ }\OtherTok{NA}

\CommentTok{# apply function to each row}
\ControlFlowTok{for}\NormalTok{ (i }\ControlFlowTok{in} \DecValTok{1}\OperatorTok{:}\KeywordTok{nrow}\NormalTok{(param_df))\{}
\NormalTok{  param_df[i,}\DecValTok{3}\NormalTok{] <-}\StringTok{ }\KeywordTok{lambda_predict}\NormalTok{(}\DataTypeTok{m=}\KeywordTok{lambda_model}\NormalTok{(),}
                                  \DataTypeTok{degree_days=}\NormalTok{param_df[i,}\DecValTok{2}\NormalTok{],}
                                  \DataTypeTok{snow=}\NormalTok{param_df[i,}\DecValTok{1}\NormalTok{])}
\NormalTok{\}}
\end{Highlighting}
\end{Shaded}

\hypertarget{problems-with-the-model}{%
\subsection{Problems with the model}\label{problems-with-the-model}}

Scanning through the completed \texttt{param\_df} data frame reveals
impossibly high values of lambda for most parameter combinations. The
problem is coming from the degree days data:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{qplot}\NormalTok{(}\DataTypeTok{x=}\NormalTok{best_preds[,}\DecValTok{1}\NormalTok{],}\DataTypeTok{y=}\NormalTok{df_clean}\OperatorTok{$}\NormalTok{Lambda)                                  }
\end{Highlighting}
\end{Shaded}

\includegraphics{BromusUpdate_files/figure-latex/unnamed-chunk-10-1.pdf}

Because of the cluster of small values at 800 degree days, the quadratic
term in the regression model has a positive coefficient, so \(\lambda\)
increases at both high and low degree days.

I don't see an easy solution, so for now, I am going to drop back to a
model just using snow cover, which has a negative coefficient for the
quadratic term and therefore decreases lambda at more extreme values.

\hypertarget{simple-snow-cover-model}{%
\subsection{Simple snow cover model}\label{simple-snow-cover-model}}

So, here are the reworked functions just using snow cover:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# Function Lambda Model2-----------------------------------------------}
\CommentTok{# inputs: lambda vector, and data frame of predictor variables}
\CommentTok{# outputs: linear model for forecasting}

\NormalTok{lambda_model2 <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(}\DataTypeTok{y=}\NormalTok{df_clean}\OperatorTok{$}\NormalTok{Lambda,}\DataTypeTok{z=}\NormalTok{best_preds)\{}
\NormalTok{  y <-}\StringTok{ }\NormalTok{df_clean}\OperatorTok{$}\NormalTok{Lambda}
  \CommentTok{# x_1 <- z[,1] # Grow.Season.DD.all.days}
\NormalTok{  x_}\DecValTok{2}\NormalTok{ <-}\StringTok{ }\NormalTok{z[,}\DecValTok{3}\NormalTok{]  }\CommentTok{# Snow.Cover}
\NormalTok{  forecast_model <-}\StringTok{ }\KeywordTok{lm}\NormalTok{(y}\OperatorTok{~}\StringTok{ }\KeywordTok{poly}\NormalTok{(x_}\DecValTok{2}\NormalTok{,}\DecValTok{2}\NormalTok{))}
\NormalTok{\}}

\CommentTok{# Function Lambda predict2 -------------------------------------------------}
\CommentTok{# inputs: linear model, degree days (vector), snow cover (vector)}
\CommentTok{# outputs: predicted lambda}

\NormalTok{lambda_predict2 <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(}\DataTypeTok{m=}\NormalTok{my_model,}\DataTypeTok{snow=}\KeywordTok{c}\NormalTok{(}\FloatTok{0.83}\NormalTok{,.}\DecValTok{74}\NormalTok{)) \{}

\NormalTok{z <-}\StringTok{ }\KeywordTok{predict}\NormalTok{(}\DataTypeTok{object=}\NormalTok{m, }\DataTypeTok{newdata=}\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{x_2=}\NormalTok{snow))}
\NormalTok{z[z}\OperatorTok{<}\DecValTok{0}\NormalTok{] <-}\StringTok{ }\DecValTok{0}   \CommentTok{# reset to 0 any predicted negative values for lambda}
\KeywordTok{return}\NormalTok{(z)}
\NormalTok{\}}

\NormalTok{param_df2 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{snow_cover=}\KeywordTok{seq}\NormalTok{(}\DataTypeTok{from=}\NormalTok{snow_low, }\DataTypeTok{to=}\NormalTok{snow_high, }\DataTypeTok{length.out=}\DecValTok{20}\NormalTok{),}
                        \DataTypeTok{lambda=}\OtherTok{NA}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Now with these simplified functions, we run the model once again:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# apply function to each row}
\ControlFlowTok{for}\NormalTok{ (i }\ControlFlowTok{in} \DecValTok{1}\OperatorTok{:}\KeywordTok{nrow}\NormalTok{(param_df2))\{}
\NormalTok{  param_df2[i,}\DecValTok{2}\NormalTok{] <-}\StringTok{ }\KeywordTok{lambda_predict2}\NormalTok{(}\DataTypeTok{m=}\KeywordTok{lambda_model2}\NormalTok{(),}
                                  \DataTypeTok{snow=}\NormalTok{param_df2[i,}\DecValTok{1}\NormalTok{])}
\NormalTok{\}}
\NormalTok{plot_}\DecValTok{1}\NormalTok{ <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(param_df2,}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x=}\NormalTok{snow_cover,}\DataTypeTok{y=}\NormalTok{lambda)) }\OperatorTok{+}
\StringTok{          }\KeywordTok{geom_point}\NormalTok{() }\OperatorTok{+}
\StringTok{          }\KeywordTok{xlab}\NormalTok{(}\StringTok{"Snow Cover"}\NormalTok{) }\OperatorTok{+}
\StringTok{          }\KeywordTok{ylab}\NormalTok{(}\StringTok{"Predicted Lambda"}\NormalTok{)}

\KeywordTok{plot}\NormalTok{(plot_}\DecValTok{1}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{BromusUpdate_files/figure-latex/unnamed-chunk-12-1.pdf}

These results look much more reasonable. The maximum predicted
\(\lambda\) is only 2.5, which is still high, but not unreasonable. And
positive population growth (\(\lambda > 1\)) is predicted for snow cover
values from \textasciitilde{} 35\% to 99\%.

And, for comparison, here is a plot of the original data from which the
prediction equation was derived:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{df <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{Snow.Cover=}\NormalTok{df_clean}\OperatorTok{$}\NormalTok{Snow.Cover, }\DataTypeTok{Lambda=}\NormalTok{df_clean}\OperatorTok{$}\NormalTok{Lambda)}
\NormalTok{plot_}\DecValTok{2}\NormalTok{ <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(df,}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x=}\NormalTok{Snow.Cover,}\DataTypeTok{y=}\NormalTok{Lambda)) }\OperatorTok{+}
\StringTok{          }\KeywordTok{geom_point}\NormalTok{() }\OperatorTok{+}
\StringTok{          }\KeywordTok{xlab}\NormalTok{(}\StringTok{"Snow Cover"}\NormalTok{) }\OperatorTok{+}
\StringTok{          }\KeywordTok{ylab}\NormalTok{(}\StringTok{"Observed Lambda"}\NormalTok{)}

\KeywordTok{plot}\NormalTok{(plot_}\DecValTok{2}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{BromusUpdate_files/figure-latex/unnamed-chunk-13-1.pdf}

\hypertarget{remaining-questions-and-issues}{%
\subsection{Remaining questions and
issues}\label{remaining-questions-and-issues}}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Obviously, the biggest problem is what to do about these environmental
  variables like degree days that do not have well-behaved quadratic
  functions. I wonder if this is an artifact caused by snow cover being
  manipulated, but environmental variables not being controlled.
  However, I thought that Alden had provided me with just data from all
  the control plots, so I don't know if this is the issue. I was really
  hoping we could have at least two predictor variables in the
  regression model, but that may not work for this data set.
\item
  In order for Matt to use this model for continental projections, we
  will need snow cover for each pixel in the continental landscape. From
  talking with Alden, it sounds like those data will be available,
  although I don't know much about the grain sizes and resolution.
\item
  In the first set of regression models, there is a strong effect of
  initial density, suggesting density dependence, but both the linear
  and quadratic coefficients are positive, so this is actually some form
  of inverse density dependence. I can't tell if this is an artifact, or
  if the density effects are what is actually driving the other patterns
  that we are seeing. But we need to discuss this more.
\item
  Alden's data set has additional columns indicating snow conditions
  (Upwind, downwind, plus, minus). I have ignored those designations in
  this analysis, but they obviously contribute to variation in
  \(\lambda\). If you want, it would be easy enough to go back and
  subscript the data to fit to fewer points. That will reduce the
  variance, although I don't think it is going to change the shape of
  the curves that much.
\item
  As you can see, the basic model with snow cover looks reasonable, but
  we need to figure out what, if anything to do with the other
  environmental variables. Once we get this sorted out and agree on a
  final regression model, I will clean up the code so that Matt has a
  single function to work with for generating maps.
\item
  Michael L. and Alden (and Jen??), we should probably talk again on
  skype after you have digested these results. I will hold off on any
  other analyses until we have talked.
\end{enumerate}

\end{document}

### 2019-07-30 Scripts/BromusUpdate30Jul2019.Rmd
Begin new update script for analyses with density and log(lambda)

--- 2019-07-20 Scripts/BromusUpdate.Rmd
@@
 ---
-title: "*Bromus* Initial Analysis"
+title: '*Bromus* Initial Analysis'
 author: "Nick Gotelli"
-date: "7/19/2019"
+date: "7/30/2019"
 output:
+  pdf_document: default
   html_document:
@@
 For a number of control and snow-manipulated plots, we have the estimated lambda $(N_{t + 1}/N_{t})$ based on annual counts of cheatgrass for 7 consecutive years.
+For incorporating density dependence, we will use $log(\lambda) = r$.
 
@@
 df_preds <- df_clean[,8:18]           # create dataframe of predictor variables
+df_clean$r <- log(df_clean$Lambda)    # create response variable r from log(lambda)
 ```
@@
 
-quad_fit <- function(x=df_clean$Snow.Cover, y=df_clean$Lambda){
+quad_fit <- function(x=df_clean$Snow.Cover, y=df_clean$r){
   M1 <- summary(lm(y~x))
@@
 # map all predictor variables to quad_fit
-summary_df <- map(df_preds,quad_fit,y=df_clean$Lambda)
+summary_df <- map(df_preds,quad_fit,y=df_clean$r)
 
@@
 A couple things jump out from these univariate regressions.
-First, the quadratic model almost always fits better than the corresponding linear model (even after imposing a penalty for the additional parameter), so we have good theoretical grounds and empirical support for going with the quadratic models.
+First, the quadratic model often fits better than the corresponding linear model (even after imposing a penalty for the additional parameter), so we have good theoretical grounds and empirical support for going with the quadratic models.
+
+The four strongest predictors were:
+
+- snow cover (quadratic r^2^ = 0.838)
+- initial density (linear r^2^ = 0.638)
+- winter precip total (quadratic r^2^ = 0.661)
+- growing season precipitation rain (quadratic r^2^ = 0.559)
+
+I had originally intended to include one of the temperature variables, but they perform poorly compared to precip and snow cover variables, so we will start with just these initial 4.
+The linear regression of r versus initial density is the relationship for a simple logistic growth model.
+The performance of the quadratic model was virtually identical, so there is no evidence for Allee effects or other more complex functions relating per capita growth rate to density.
+So we will stick with the linear model.
 
@@
 ```{r}
+str(df_preds)
 # best variables are:
-# Grow.Season.DD.all.days (4)
-# Grow.Season.Precip.Total (5)
-# Snow.Cover (6)
-best_preds <- df_preds[,4:6]
+# snow.cover (6)
+# density (1)
+# winter precip total. (11) 
+# grow season precip rain (3)
 
-y <- df_clean$Lambda
+best_preds <- df_preds[,c(6,1,11,3)]
+
+y <- df_clean$r
 x_1 <- best_preds[,1]
@@
 x_3 <- best_preds[,3]
+x_4 <- best_preds[,4]
 
-model_1 <- summary(lm(y~poly(x_1,2) + poly(x_2,2) + poly(x_3,2)))
+final_df <- data.frame(r=y,
+                       snow_cover=x_1,
+                       density=x_2,
+                       win_precip <- x_3,
+                       gs_rain <- x_4)
+
+model_dd <- summary(lm(data=final_df,r~density))
+print(model_dd)
+model_dd <- summary(lm(data=final_df,r~poly(density,2)))
+print(model_dd)
+ggplot(data=final_df,aes(x=density,y=r)) +
+  geom_point() +
+  stat_smooth(method="lm", se=TRUE, fill=NA,
+                formula=y ~ poly(x, 2, raw=TRUE),colour="red") +
+  stat_smooth(method="lm", se=TRUE, fill=NA,
+                formula=y ~ poly(x, 1, raw=TRUE),colour="blue")
+
+model_1 <- summary(lm(y~poly(x_1,2) + x_2 + poly(x_3,2) + poly(x_4,2)))
 print(model_1)
@@
 
-lambda_model <- function(y=df_clean$Lambda,z=best_preds){
-  y <- df_clean$Lambda
+lambda_model <- function(y=df_clean$r,z=best_preds){
+  y <- df_clean$r
   x_1 <- z[,1] # Grow.Season.DD.all.days
@@
                         degree_days=seq(from=dd_low, to=dd_high,length.out=20))
-param_df$lambda <- NA
+param_df$r <- NA
 
@@
 ```{r}
-qplot(x=best_preds[,1],y=df_clean$Lambda)
+qplot(x=best_preds[,1],y=df_clean$r)
 
@@
 
-lambda_model2 <- function(y=df_clean$Lambda,z=best_preds){
-  y <- df_clean$Lambda
+lambda_model2 <- function(y=df_clean$r,z=best_preds){
+  y <- df_clean$r
   # x_1 <- z[,1] # Grow.Season.DD.all.days
@@
 ```{r}
-df <- data.frame(Snow.Cover=df_clean$Snow.Cover, Lambda=df_clean$Lambda)
+df <- data.frame(Snow.Cover=df_clean$Snow.Cover, Lambda=df_clean$r)
 plot_2 <- ggplot(df,aes(x=Snow.Cover,y=Lambda)) +

### 2019-09-04 Scripts/MatrixFunctions.R
Bromus updates and medfly clean up

@@ -40,7 +40,7 @@ library(parallel)
 # input: x=vector of temperatures, y=vector of transition elements
 # output: fitted regression model
 est_trans <- function(x=runif(4),y=runif(4)){
-	fitted <- lm(y~poly(x,2))
+	fitted <- lm(y~poly(x,2,raw=TRUE))
 	return(fitted)
 	
 }
@@ -54,7 +54,7 @@ pred_trans <- function(new=runif(1,0,40),f=NULL){
             	x <- runif(4)
             	y <- runif(4)
             	df<- data.frame(x,y)   
-            	f<- lm(y~poly(x,2),data=df)}
+            	f<- lm(y~poly(x,2,raw=TRUE),data=df)}
            
 	z <- predict(object=f, newdata=data.frame(x=new))
 	return(z)

### 2019-09-04 Scripts/MattsGraphicsCode.R

@@ -66,13 +66,15 @@ lambda_rasts <- mclapply(1:12, function(x, tempRasts){
   rast <- tempRasts[[x]]
   keep <- !is.na(rast[]) # index NAs
 #-------------------------------------
-  lamb <- lam_gen(rast[keep]) # calc lamba for each temp
-  lamb[lamb<1] <-0 # clip non-increasing values to 0
-#-------------------------------------
-  # lamb <- lam2_gen(x=exp_tmp,
-  # 								 y=emp_lam,
-  # 								 z=rast[keep]) # calc lamba for each temp
+  # original function based on transition matrices
+  # lamb <- lam_gen(rast[keep]) # calc lamba for each temp
   # lamb[lamb<1] <-0 # clip non-increasing values to 0
+#-------------------------------------
+# use lambda2_gen, which is based on lambdas, not transition matrices
+  lamb <- lam2_gen(x=exp_tmp,
+  								 y=emp_lam,
+  								 z=rast[keep]) # calc lamba for each temp
+  lamb[lamb<1] <-0 # clip non-increasing values to 0
 #-------------------------------------
   rast[which(keep)] <- unlist(lamb) # assign lamba back to raster
   return(rast)}, tempRasts=tmean, mc.cores=2)

### 2019-09-04 Scripts/BromusUpdate0Sep2019.Rmd
Update Bromus model with new data and final regression stats

---
title: '*Bromus* Density Dependence Analysis'
author: "Nick Gotelli"
date: "4 September 2019"
output:
  pdf_document: default
  html_document:
    highlight: tango
    theme: cerulean
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Updatre status
This update uses the expanded data set from Alden and Catherine and cleans up the coefficients in the quadratic function

## Data sources and cleaning
These data were provided by Alden Griffith and Michael Loic. After talking with Alden, we decided to use $\lambda$ values that had been estimated from changes in the number of individuals in marked plots, rather than analysis of stage transition matrices. For a number of control plots, Alden and Michael have estimated lambda $(N_{t + 1}/N_{t})$ based on annual counts of cheatgrass for two consecutive years. For some of the plots, there are actually 7 consecutive years of data, but I am ignoring plot identity, and simply using $\lambda$ calculated for pair of consecutive abundances. For each of these measures, the variable `Starting.Density` gives $N_t$, the abundance in the first year.

For a simple logistic growth equation, there should be a negative linear relationship between  $r =        log(\lambda)$ and $N_t$.

I cleaned the initial data file (`Bromus Lambda and Climate Data.csv`), moving a notes column to initial comment lines, and adding 4 missing values as `NA`. The modified data are in `CleanBromusData.csv`.

```{r}
# Preliminaries -----------------------------------------------------------
library(tidyverse)

# Get input data ----------------------------------------------------------
df <- read.table(file="../CleanData/Bromus_Clean_V2_Aug-28-19.csv",
                 sep=",",header=TRUE,stringsAsFactors=FALSE)
df <- df[,-ncol(df)]                 # remove empty final column
df_clean <- df[complete.cases(df),]  # remove 4 missing lambda

df_preds <- df_clean[,8:18]           # create dataframe of predictor variables
df_clean$r <- log(df_clean$Lambda)    # create response variable r from log(lambda)
```

## Initial regression models
With the data frame set up, the first analysis was to use each of the predictor variables (snow, degree-days, precipitation, etc.), with $r$ as the response variable. For each variable, I built a linear and a quadratic regression model.

```{r}
# Function Model Fit ------------------------------------------------------
# inputs: response variable, predictor variable
# outputs: adjusted r2 for linear and quadratic regression models

quad_fit <- function(x=df_clean$Snow.Cover, y=df_clean$r){
  M1 <- summary(lm(y~x))
  M2 <- summary(lm(y~poly(x,2,raw=TRUE)))
  return(c(M1$adj.r.squared,M2$adj.r.squared))
}
# -------------------------------------------------------------------------
# map all predictor variables to quad_fit
summary_df <- map(df_preds,quad_fit,y=df_clean$r)

# put into a summary data frame and add column labels
summary_df <- t(bind_rows(summary_df))
colnames(summary_df) <- c("Linear","Quadratic")
# -------------------------------------------------------------------------
```

## First screening of predictor variables
In this table, each row is a different model with a single predictor variable, and the data frame shows the adjusted r^2^ for each model. The two columns are for a linear model $(y \sim x)$ and a quadratic model $(y \sim x + x^2)$.

```{r}
print(summary_df)
```

The quadratic model often fits better than the corresponding linear model (even after imposing a penalty for the additional parameter), so we have good theoretical grounds and empirical support for going with the quadratic models. However, the quadratic model does not provide much improvement for the density dependence, so we will go with the linear model, which corresponds to a plain-vanilla logistic growth equation.

The strongest predictors were:

- snow cover (quadratic r^2^ = 0.759)
- winter precip snow (quadratic r^2^ = 0.612)
- winter precip total (quadratic r^2^ = 0.555)
- winter precip rain (quadratic r^2^ = 0.509)
- growing season precip rain (quadratic r^2^ = 0.469)
- initial density (linear r^2^ = 0.368)

I had originally intended to include one of the temperature variables, but they perform poorly compared to precip and snow cover variables, so we will start with just these initial 4. Many of these snow variables will be intercorrelated, so we should be able to shrink down to a smaller set of predictors. These results are qualitatively similar to what we found with the initial data set, but the density dependence is weaker in this expanded data set.

## Simple linear density dependence model

We start by ignoring all the climate variables and just fit the collection of annual $\lambda$ values to the linear function we expect for density dependence in a plain-vanilla logistic growth model:

```{r}
# best variables are:
# snow.cover (6)
# winter precip snow (7)
# winter precip total (11)
# winter precip rain (8)
# growing season precip rain (3)
# density (1)
best_preds <- df_preds[,c(6,7,11,8,3,1)]

y <- df_clean$r
x_1 <- best_preds[,1]
x_2 <- best_preds[,2]
x_3 <- best_preds[,3]
x_4 <- best_preds[,4]
x_5 <- best_preds[,5]
x_6 <- best_preds[,6]

final_df <- data.frame(r=y,
                       snow_cover=x_1,
                       win_precip_snow=x_2,
                       win_precip_tot=x_3,
                       win_precip_rain=x_4,
                       grow_seas_precip_rain=x_5,
                       density=x_6)

model_dd <- summary(lm(data=final_df,r~density))
print(model_dd)
ggplot(data=final_df,aes(x=density,y=r)) +
  geom_point() +
  stat_smooth(method="lm", se=TRUE, fill=NA,
                formula=y ~ poly(x, 1, raw=TRUE),colour="red") 
```

These results look pretty nice. The relationship is roughly linear, and we pick up a sizeable chunk of the variation in $r$. The intercept is 0.642016, which corresponds to the maximum $r$ that occurs with no crowding ($N_t$ = 0). This is the constant $r$ that appears in the logistic growth equation. It is large ($\lambda = exp(r) = 1.9$), but reasonable for an annual plant. You can see there are some low-density plots that have even higher values than this.

The other parameter of interest is to set $r = 0$ and solve for $N$. This is the estimate of the carrying capacity $K$ in the logistic growth equation. For these data, we get K = 484 individuals.

Alden and Michael, do these values of $r = 1.64 individuals/individual\times year$ and $K = 484 individuals$ seem biologically reasonable for *Bromus**?

## Quadratic model for snow cover

```{r}
model_snow_cover <- summary(lm(data=final_df,r~poly(snow_cover,2)))
print(model_snow_cover)
ggplot(data=final_df,aes(x=snow_cover,y=r)) +
  geom_point() +
  stat_smooth(method="lm", se=TRUE, fill=NA,
                formula=y ~ poly(x, 2, raw=TRUE),colour="red") 
```

## Quadratic model for winter precipitation snow

```{r}
model_win_precip <- summary(lm(data=final_df,r~poly(win_precip_snow,2)))
print(model_win_precip)
ggplot(data=final_df,aes(x=win_precip_snow,y=r)) +
  geom_point() +
  stat_smooth(method="lm", se=TRUE, fill=NA,
                formula=y ~ poly(x, 2, raw=TRUE),colour="red") 
```

## Quadratic model for winter precip total

```{r}
model_win_precip_tot <- summary(lm(data=final_df,r~poly(win_precip_tot,2)))
print(model_win_precip_tot)
ggplot(data=final_df,aes(x=win_precip_tot,y=r)) +
  geom_point() +
  stat_smooth(method="lm", se=TRUE, fill=NA,
                formula=y ~ poly(x, 2, raw=TRUE),colour="red") 
```

## Quadratic model for winter precip rain

```{r}
model_win_precip_rain <- summary(lm(data=final_df,r~poly(win_precip_rain,2)))
print(model_win_precip_rain)
ggplot(data=final_df,aes(x=win_precip_rain,y=r)) +
  geom_point() +
  stat_smooth(method="lm", se=TRUE, fill=NA,
                formula=y ~ poly(x, 2, raw=TRUE),colour="red") 
```

## Quadratic model for growing season precip rain

```{r}
model_grow_seas_precip_rain <- summary(lm(data=final_df,r~poly(grow_seas_precip_rain,2)))
print(model_grow_seas_precip_rain)
ggplot(data=final_df,aes(x=grow_seas_precip_rain,y=r)) +
  geom_point() +
  stat_smooth(method="lm", se=TRUE, fill=NA,
                formula=y ~ poly(x, 2, raw=TRUE),colour="red") 
```

The quadratic models all look good, except this one for growing season rain, which has a positive quadratic term.

## Saturated regression model with all predictors

```{r}
model_saturated <- summary(lm(data=final_df,r~poly(snow_cover,2, raw=TRUE) +
                                              poly(win_precip_snow,2, raw=TRUE) +
                                              poly(win_precip_tot,2, raw=TRUE) +
                                              poly(win_precip_rain,2, raw=TRUE) +
                                              poly(grow_seas_precip_rain,2, raw=TRUE) +
                                              poly(density,1)))

print(model_saturated)
```

## Reduced model

Based on these initial results, we can winnow out a few variables to start with:

```{r}
model_reduced <- summary(lm(data=final_df,r~poly(snow_cover,2, raw=TRUE) +
                                              poly(win_precip_snow,2, raw=TRUE) +
                                              poly(win_precip_tot,2, raw=TRUE) +
                                              poly(density,1)))

print(model_reduced)
```

## Reduced model #2

For this model, we will try dropping out `grow_seas_precip_rain` simply because that variable has a positive quadratic term, which is not biologically realistic, and it only adds about 2% variance explained.

```{r}
model_reduced2 <- summary(lm(data=final_df,r~poly(snow_cover,2, raw=TRUE) +
                                              poly(win_precip_rain,2, raw=TRUE) +
                                              poly(density,1)))

print(model_reduced2)
```

This reduced model has only 3 variables: snow cover, winter rain, and initial density, and accounts for almost 80% of the variation in population growth rate. That's about as good as it gets for regression models with observational data.

The next step will be to build a prediction function for Matt, who will then need to set up data layers for snow cover and winter rain. In the prediction model, we will set density to 0 because we are looking at colonization potential of *Bromus* in an empty continental landscape.

### 2019-09-04 Scripts/UpdatedBromusRead.R

@@ -1,0 +1,10 @@
+# Script for opening and cleaning new Bromus data
+# 04 Sept 2019
+# NJG
+
+# Read in data ------------------------------------------------------------
+
+z <- read.table("CleanData/Bromus_Clean_V2_Aug-28-19.csv",sep=",",
+                stringsAsFactors=FALSE, header=TRUE)
+str(z)
+

### 2019-09-15 Scripts/BromusRollingUpdateFile.Rmd
Test for interaction term

--- 2019-09-04 Scripts/BromusUpdate0Sep2019.Rmd
@@
 
-## Updatre status
+## 15 September 2019
+This update adds raw to all poly statements, tries adding an interaction term for snow and precip, and constructs heat maps for visualizing the Hutchinsonian niche.
+
+## Update status
 This update uses the expanded data set from Alden and Catherine and cleans up the coefficients in the quadratic function
@@
                                               poly(grow_seas_precip_rain,2, raw=TRUE) +
-                                              poly(density,1)))
+                                              poly(density,1, raw=TRUE)))
 
@@
                                               poly(win_precip_tot,2, raw=TRUE) +
-                                              poly(density,1)))
+                                              poly(density,1, raw=TRUE)))
 
@@
                                               poly(win_precip_rain,2, raw=TRUE) +
-                                              poly(density,1)))
+                                              poly(density,1, raw=TRUE)))
 
@@
 In the prediction model, we will set density to 0 because we are looking at colonization potential of *Bromus* in an empty continental landscape.
+
+## Reduced model with interaction term
+
+For this model, we will try an interaction between the linear components of snow cover and rain
+
+```{r}
+model_reduced3 <- summary(lm(data=final_df,r~poly(snow_cover,2, raw=TRUE) +
+                                              poly(win_precip_rain,2, raw=TRUE) +
+                                              poly(density,1, raw=TRUE) + 
+                                              snow_cover:win_precip_rain))
+
+print(model_reduced3)
+```
+
+Since the interaction term doesn't add anything, we will not include it and stick with the simple reduced model.
+
+## Heat map for visualizing lambda values

### 2019-09-17 BromusLambdaFunction.R
Finish script for forecasting function

@@ -1,0 +1,107 @@
+## ----setup, include=FALSE------------------------------------------------
+knitr::opts_chunk$set(echo = TRUE)
+
+
+## ------------------------------------------------------------------------
+# Preliminaries -----------------------------------------------------------
+library(tidyverse)
+
+# Get input data ----------------------------------------------------------
+
+# adjust path for grabbing input file
+df <- read.table(file="CleanData/Bromus_Clean_V2_Aug-28-19.csv",
+                 sep=",",header=TRUE,stringsAsFactors=FALSE)
+df <- df[,-ncol(df)]                 # remove empty final column
+df_clean <- df[complete.cases(df),]  # remove 4 missing lambda
+
+df_preds <- df_clean[,8:18]           # create dataframe of predictor variables
+df_clean$r <- log(df_clean$Lambda)    # create response variable r from log(lambda)
+
+
+
+## ------------------------------------------------------------------------
+# best variables are:
+# snow.cover (6)
+# winter precip snow (7)
+# winter precip total (11)
+# winter precip rain (8)
+# growing season precip rain (3)
+# density (1)
+best_preds <- df_preds[,c(6,7,11,8,3,1)]
+  
+y <- df_clean$r
+x_1 <- best_preds[,1]
+x_2 <- best_preds[,2]
+x_3 <- best_preds[,3]
+x_4 <- best_preds[,4]
+x_5 <- best_preds[,5]
+x_6 <- best_preds[,6]
+
+final_df <- data.frame(r=y,
+                       snow_cover=x_1,
+                       win_precip_snow=x_2,
+                       win_precip_tot=x_3,
+                       win_precip_rain=x_4,
+                       grow_seas_precip_rain=x_5,
+                       density=x_6)
+
+
+
+
+
+
+## ------------------------------------------------------------------------
+model_reduced2 <- lm(data=final_df,r~poly(snow_cover,2, raw=TRUE) +
+                                              poly(win_precip_rain,2, raw=TRUE) +
+                                              poly(density,1, raw=TRUE)) 
+                                              
+## ------------------------------------------------------------------------
+#-------------------------------------
+# FUNCTION pred_brom_lam
+# predict lambda based on snow cover, precip, and density
+# input: f=fitted model, new= data frame for prediction
+# output: new model coefficient
+pred_brom_lam <- function(new=NULL,f=model_reduced2){
[…]

### 2019-09-17 Scripts/Bromus_Lambda_Final.R

@@ -1,0 +1,5 @@
+# This script sets up the final best-fitting model for predicting
+# Bromus population growth as a function of snow cover, winter rain precipitation,
+# and density (which is held constant at 0).
+
+# The model uses an updated data file provided by Alden Griffith and Michael Loic

### 2019-09-19 BromusLambdaFunction.R
Calculate K and associated stats and heat map

@@ -105,3 +105,31 @@ str(hutchinson_df)
 # heat_map <- ggplot(data=heat_map_data,mapping=aes(x=snow_cover,y=win_precip_rain,fill=r))
 # heat_map + geom_tile() + scale_fill_viridis_c()
 
+
+# Code for estimating carrying capacity -----------------------------------
+
+# cf <- summary(model_reduced2)$coefficients[,1]    # grab the coefficients
+# d <- heat_map_data # rename the data frame
+
+# now create new column with K values
+# heat_map_data$K <- (cf[1] + cf[2]*d[[1]] + cf[3]*d[[1]]^2 +
+#                    cf[4]*d[[2]] + cf[5]*d[[2]]^2)/(-1*cf[6])
+
+# convert negatives to 0
+# heat_map_data$K[heat_map_data$K<=0] <- 0
+
+# create heat map
+# heat_map_k <- ggplot(data=heat_map_data,mapping=aes(x=snow_cover,y=win_precip_rain,fill=K))
+# heat_map_k + geom_tile() + scale_fill_viridis_c()
+
+# summarize distribution of K values greater than 0
+# real_k <- heat_map_data$K[heat_map_data$K>0]
+# summary(real_k)
+
+# get snow and precip for smallest positive K
+# heat_map_data$snow_cover[which(heat_map_data$K==min(heat_map_data$K[heat_map_data$K>0]))]
+# heat_map_data$win_precip_rain[which(heat_map_data$K==min(heat_map_data$K[heat_map_data$K>0]))]
+
+# get snow and precip for largest K
+# heat_map_data$snow_cover[which(heat_map_data$K==max(heat_map_data$K))]
+# heat_map_data$win_precip_rain[which(heat_map_data$K==max(heat_map_data$K))]

### 2020-06-23 ManuscriptDraft/ManuscriptDraft.Rmd
add manuscript draft

---
title: "Invasion Demography Draft"
author: "Gotelli et al."
date: "6/23/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rprojroot::find_rstudio_root_file()
```

A major research focus in ecology is forecasting how species and communities will change in response to climate warming and other environmental drivers (). Such shifts can potentially bring together new combinations of species that previously have not co-ccurred before, and the geograpnic ranges of many non-native species do seem to be changing recent decades. Some of these species have strong direct effects on native assemblages as predators, competitors, and parasites, as well as indirect cascading effects through habitat modifications.

Although a variety of forecasting models are currently used to predict changes in geographic ranges, most of them rely on three basic data layers. The first is a set of geographic coordinates that specify the locations of current populations of the species. These records come from gbif, inaturalist, and other sources of museum records and private surveys in which the presence of species has been recorded. The second data layer is a set of environmental variable and climate variables, especially measures of temperature and precipitation, that have been measured, interpolated, and then downscaled for the locations of the individual populations. These two data layers are analyzed with a variety of species distribution models (SDMs), including MaxEnt, and more recent neural network and random forest approaches. These algorithms seek to estimate and optimize the probability of species occurrence as a function of environmental variables. The fitted model can then be used to forecast the probability of occurrence (or habitat suitability) at any location on a (terrestrial map) as a function of contemporary climate variables in that location. The third data layer is a forecast of future environmental conditions, often derived from climate change models, at different locations. This forecasting layer for future conditions can be combined with the species distribution model that was fit with contemporary climate variables to estimate suitability of future conditions for species occurrences.

In spite of the diversity of methods for analyzing data of this form, they all suffer from two important limitations. The first limitation is that these models assume that populations are in equilibrium with their environmental conditions. In this equilibrium state, population occurrences would represent locations that are suitable for populations to persist. However, the pace of climate change has been rapidly increasing in recent decades, and even short-lived species may no longer be in equilibrium with their current abiotic environment. Indeed, many recent studies have emphasized that rates of spread of species may not be able to keep pace with the projected "climate velocity, and many models of climatic debt emphasize a disequilibrium between the current composition of a community, and the composition that would be predicted assuming an equilibrium with climate.

The second limitation is that most of the data sets used in species distribution modeling are based on presence-only data. In the absence of widespread standardized surveys, we have only the location of the presences of a species, but not locations in which absences have been recorded. Without true absences, most algorithms impute "pseudo-absences" typically by randomly placing absence points within an arbitrarily bounded domain that encompasses the recorded species occurrences. But for many groups of organisms (especially plants and invertebrates), many of the locations were pseudoabsences are imputed probably contain the species, but have not been properly sampled. The location of these pseudoabsences has a very large affect on the outcome of the model, but often the domain boundaries provided by users are arbitrary and can simply reflect the original raster that contained the occurrence records.

These limitations have made the forecasting of both native and non-native species highly uncertain.  In this paper, we propose an alternative framework for forecasting the distribution of non-native species, that does not depend on the spatial pattern of population presences and absences, does not assume that populations are in equilibrium with their environment, and does not assume that the presence of a population is a good indicator of the climatic niche of a species.

Instead, we argue for species distribution modeling that is based on demographic data. Basic life-table data describe rates of birth, individual growth, and death as a function of the age or size or stage of the organism. These data can be collected in a variety of ways, and can be converted to an estimate of r. In a model of exponential population growth, r is the intrinsic rate of increase, measured in units of time/individual*individual. Alternatively, when demographic rate data are not available, r can be estimated from sequential population censuses as loge(N(t+1)/N(t)). With either life table or population growth data, r is a natural metric for the potential for population persistence in different environmental conditions. If r is greater than 0.0, instantaneous birth rates exceed instantaneous death rates, and the population is able, at least in the short-term, to persist and grow. If r is less than 0.0, instantaneous death rates exceed instantaneous birth rates, and the population will, in the long-run, decline to extinction.

Here, we use examples from laboratory studies, field experiments, and field observational studies to illustrate three case studies of invasive species for which we forecast population growth rates across a map of the continental U.S. In each case, we find that the models generate realistic maps of habitat suitability when fitted with contemporary climate layers. These predictions based on demographic data are either comparable to, or better than, published predictions based on traditional species distribution models.

METHODS

The general framework for these analyses is simplified and adapted from Gotelli amd Ellison (2002, 2006). Three general equations describe the relationship between demography, environmental variables, and geographic location:

1) r = f(E)

The function f establishes the relationship between population growth rate (lambda) and the environmental conditions (E). In an experimental study, a life table response experiment (Caswell 200x)  would be used to rear organisms under constant conditions. Each level of temperature or some other continuous environmental variable would represent a separate, replicated treatment.  Alternatively, for a field based population study, a researcher would measure population size at two consecutive time steps, and the associated environmental conditions over that same interval. For both the experimental and non-experimental data, the fitted function f describes the relationship between environmental conditions and lambda. For an environmental variable z, we propose the simplest model for predicting r=log(lambda) should be

f(E) = a + bz + cz^2

In a least squares regression model, this is a quadratic function with fitted constants a, b, and c. We favor this model for theoretical reasons: for any environmental variable, there should be an optimal, intermediate level of the variable that maximizes population growth rate. Population growth rate decreases and the environmental variable moves away from the optimum. If there is a range of values of z within which population growth rate is positive, this range will include the peak. Although we have presented this framework in the context of a single environmental driver, the model can be easily extended with multiple regression to incorporate 2 or more drivers, such as precipitation and temperature.

2) E = g(x,y)

The function g describes the relationship between the environmental variable (E) and the location of a point in the landscape, based on its geographic coordinates (x,y). This data layer is usually created from smoothed and interpolated weather data, and is rarely modeled by the investigator. The data layer may be available for past, current, or future conditions based on climate forecasting and hind-casting models.

3) r(xy) = f(g(xy))

The final step is to estimate r for each pixel in the landscape by combining the calculations from (1) and (2). r can be treated as a continuous variable representing the ability of the population to achieve short-term population growth, given the set of environmental conditions at location (x,y). r is an unbounded continuous variable, and the more positive the value of r, the higher the growth rate, and presumably the better the environmental conditions are for population growth. A critical value in this continuum is r = 0, which is the dividing line between increasing and decreasing population growth rate. Modeling r as a function of environmental variables is an empirical realization of Hutchinson's definition of the niche: an n-dimensional hypervolume describing a combination of variables for which dN/dt > 0.

### 2020-06-23 ManuscriptDraft/ManuscriptDraft.Rmd
clean up introductory text and equations

@@
 
+## Introduction
+
 A major research focus in ecology is forecasting how species and communities will change in response to climate warming and other environmental drivers ().
-Such shifts can potentially bring together new combinations of species that previously have not co-ccurred before, and the geograpnic ranges of many non-native species do seem to be changing recent decades.
-Some of these species have strong direct effects on native assemblages as predators, competitors, and parasites, as well as indirect cascading effects through habitat modifications.
+Such shifts can potentially bring together new combinations of species that previously have not co-ccurred before (); the geograpnic ranges of many non-native species do seem to be changing recent decades.
+Some of these species have strong direct effects on native assemblages as predators (), competitors (), and parasites (), as well as indirect cascading effects through habitat modifications ().
 
-Although a variety of forecasting models are currently used to predict changes in geographic ranges, most of them rely on three basic data layers.
+Although a variety of forecasting models are currently used to predict changes in geographic ranges (), most of them rely on three basic data layers.
 The first is a set of geographic coordinates that specify the locations of current populations of the species.
-These records come from gbif, inaturalist, and other sources of museum records and private surveys in which the presence of species has been recorded.
+These records come from data portals such as GBIF (), iNaturalist (), and other sources of museum records and individual surveys in which the presence of species has been recorded.
+
 The second data layer is a set of environmental variable and climate variables, especially measures of temperature and precipitation, that have been measured, interpolated, and then downscaled for the locations of the individual populations.
-These two data layers are analyzed with a variety of species distribution models (SDMs), including MaxEnt, and more recent neural network and random forest approaches.
-These algorithms seek to estimate and optimize the probability of species occurrence as a function of environmental variables.
+These first two data layers can be analyzed with a variety of species distribution models (SDMs), including MaxEnt (), and more recent neural network () and random forest models ().
+These algorithms seek to estimate and optimize the suitability of the habitat in each pixel of the landscape as a function of environmental variables.
 The fitted model can then be used to forecast the probability of occurrence (or habitat suitability) at any location on a (terrestrial map) as a function of contemporary climate variables in that location.
-The third data layer is a forecast of future environmental conditions, often derived from climate change models, at different locations.
-This forecasting layer for future conditions can be combined with the species distribution model that was fit with contemporary climate variables to estimate suitability of future conditions for species occurrences.
+
+The third data layer is a forecast of future environmental conditions, often derived from climate change models (), at different locations.
+This forecasting layer for future conditions can be combined with the species distribution model that was fit with contemporary climate variables to estimate suitability of future conditions for species occurrences ().
 
 In spite of the diversity of methods for analyzing data of this form, they all suffer from two important limitations.
-The first limitation is that these models assume that populations are in equilibrium with their environmental conditions.
-In this equilibrium state, population occurrences would represent locations that are suitable for populations to persist.
-However, the pace of climate change has been rapidly increasing in recent decades, and even short-lived species may no longer be in equilibrium with their current abiotic environment.
-Indeed, many recent studies have emphasized that rates of spread of species may not be able to keep pace with the projected "climate velocity, and many models of climatic debt emphasize a disequilibrium between the current composition of a community, and the composition that would be predicted assuming an equilibrium with climate.
+The first limitation is that these models assume that populations are in equilibrium with their environmental conditions ().
+In this equilibrium state, contemporary population occurrences would represent locations that are suitable for populations to persist.
+However, the pace of climate change has accelerated in recent decades (), and even short-lived species may no longer be in equilibrium with their current abiotic environment.
+Indeed, many recent studies have emphasized that rates of spread of species may not be able to keep pace with the projected "climate velocity", and many models of climatic debt () emphasize a disequilibrium between the current composition of a community, and the composition that would be predicted assuming an equilibrium with climate ().
 
-The second limitation is that most of the data sets used in species distribution modeling are based on presence-only data.
-In the absence of widespread standardized surveys, we have only the location of the presences of a species, but not locations in which absences have been recorded.
-Without true absences, most algorithms impute "pseudo-absences" typically by randomly placing absence points within an arbitrarily bounded domain that encompasses the recorded species occurrences.
-But for many groups of organisms (especially plants and invertebrates), many of the locations were pseudoabsences are imputed probably contain the species, but have not been properly sampled.
-The location of these pseudoabsences has a very large affect on the outcome of the model, but often the domain boundaries provided by users are arbitrary and can simply reflect the original raster that contained the occurrence records.
+The second limitation is that most species distribution modeling is based on presence-only data.
+In the absence of widespread standardized surveys, we have only the location of the occurrences of a species, but not locations in which absences have been explicitly recorded.
+Without true absences, most algorithms impute "pseudo-absences" by randomly placing absence points (10,000 in the default settings for MaxEnt) uniformly within an arbitrarily bounded domain that encompasses the recorded species occurrences.
+But for many groups of organisms (especially plants and invertebrates), many of the locations where pseudoabsences are imputed probably contain the species, but have not been properly sampled.
+The location of these pseudoabsences has a very large affect on the outcome of the model (), but often the domain boundaries provided by users are arbitrary and can simply reflect the original raster that contained the occurrence records.
 
-These limitations have made the forecasting of both native and non-native species highly uncertain.
+These limitations have made the forecasting of both native and non-native species highly uncertain ().
 In this paper, we propose an alternative framework for forecasting the distribution of non-native species, that does not depend on the spatial pattern of population presences and absences, does not assume that populations are in equilibrium with their environment, and does not assume that the presence of a population is a good indicator of the climatic niche of a species.
 
-Instead, we argue for species distribution modeling that is based on demographic data.
-Basic life-table data describe rates of birth, individual growth, and death as a function of the age or size or stage of the organism.
-These data can be collected in a variety of ways, and can be converted to an estimate of r.
+Instead, we argue for species distribution modeling that is based on estimates of short-term population growth rates derived from demograpic data ().
+Basic life-table data describe rates of birth, individual growth, and death as a function of the age or size or stage of the organism ().
+These data can be collected in a variety of ways, and can be converted to an estimate of r ().
 In a model of exponential population growth, r is the intrinsic rate of increase, measured in units of time/individual*individual.
+
 Alternatively, when demographic rate data are not available, r can be estimated from sequential population censuses as loge(N(t+1)/N(t)).
@@
 
-Here, we use examples from laboratory studies, field experiments, and field observational studies to illustrate three case studies of invasive species for which we forecast population growth rates across a map of the continental U.S.
+Here, we use examples from laboratory studies, field experiments, and field observational studies to illustrate three case studies of invasive species for which we forecast population growth rates at sites across a map of the continental U.S.
+These analyses are based only on demographic data, and do not use species occurrence records to construct a forecast.
+
 In each case, we find that the models generate realistic maps of habitat suitability when fitted with contemporary climate layers.
@@
 
-METHODS
+## Methods
 
@@
 
-1) r = f(E)
+\begin{align}
+r = f(E)
+\end{align}
 
-The function f establishes the relationship between population growth rate (lambda) and the environmental conditions (E).
+The function $f$ establishes the relationship between population growth rate $(r)$ and the environmental conditions $(E)$.
 In an experimental study, a life table response experiment (Caswell 200x)  would be used to rear organisms under constant conditions.
 Each level of temperature or some other continuous environmental variable would represent a separate, replicated treatment.
+
 Alternatively, for a field based population study, a researcher would measure population size at two consecutive time steps, and the associated environmental conditions over that same interval.
-For both the experimental and non-experimental data, the fitted function f describes the relationship between environmental conditions and lambda.
-For an environmental variable z, we propose the simplest model for predicting r=log(lambda) should be
+For both the experimental and non-experimental data, the fitted function $f$ describes the relationship between $E$ and $r$.
+For an environmental variable $z$, we propose the simplest model for predicting $r$ should be
 
-f(E) = a + bz + cz^2
+$$f(E) = a + bz + cz^2$$
 
-In a least squares regression model, this is a quadratic function with fitted constants a, b, and c.
-We favor this model for theoretical reasons: for any environmental variable, there should be an optimal, intermediate level of the variable that maximizes population growth rate.
-Population growth rate decreases and the environmental variable moves away from the optimum.
-If there is a range of values of z within which population growth rate is positive, this range will include the peak.
+In a least squares regression model, this is a quadratic function with fitted constants $a$, $b$, and $c$.
+We favor this model for theoretical reasons: for any environmental variable, there should be an optimal, intermediate level of $z$ that maximizes population growth rate.
+Population growth rate decreases as the environmental variable moves away from the optimal value.
+If there is a range of values of $z$ within which $r$ is positive, this range will include the peak.
 Although we have presented this framework in the context of a single environmental driver, the model can be easily extended with multiple regression to incorporate 2 or more drivers, such as precipitation and temperature.
 
-2) E = g(x,y)
+\begin{align}
+E = g(x,y)
+\end{align}
 
-The function g describes the relationship between the environmental variable (E) and the location of a point in the landscape, based on its geographic coordinates (x,y).
-This data layer is usually created from smoothed and interpolated weather data, and is rarely modeled by the investigator.
-The data layer may be available for past, current, or future conditions based on climate forecasting and hind-casting models.
+The function $g$ describes the relationship between the environmental variable $(E)$ and the location of a point in the landscape, specified by its geographic coordinates $(x,y)$.
+This data layer is usually created from smoothed and interpolated climate data (), and is rarely modeled by the investigator.
+The data layer may be available for past, current, or future conditions based on climate forecasting and hind-casting models ().
 
-3) r(xy) = f(g(xy))
+\begin{align}
+r_{xy} = f(g(xy))
+\end{align}
 
 The final step is to estimate r for each pixel in the landscape by combining the calculations from (1) and (2).
-r can be treated as a continuous variable representing the ability of the population to achieve short-term population growth, given the set of environmental conditions at location (x,y).
-r is an unbounded continuous variable, and the more positive the value of r, the higher the growth rate, and presumably the better the environmental conditions are for population growth.
-A critical value in this continuum is r = 0, which is the dividing line between increasing and decreasing population growth rate.
-Modeling r as a function of environmental variables is an empirical realization of Hutchinson's definition of the niche: an n-dimensional hypervolume describing a combination of variables for which dN/dt > 0.
+r can be interpreted as a continuous variable representing the ability of the population to achieve short-term population growth, given the set of environmental conditions at location $(x,y)$.
+$r$ is an unbounded continuous variable, and the more positive the value of $r$, the higher the growth rate, and presumably the better the environmental conditions are for population growth.
+A critical value in this continuum is $r = 0$, which is the dividing line between increasing and decreasing population growth rate.
+Modeling r as a function of environmental variables is an empirical realization of Hutchinson's () definition of the niche: an $n$-dimensional hypervolume describing a combination of variables for which $dN/dt > 0$.

### 2020-06-28 ManuscriptDraft/ManuscriptDraft.Rmd
update raster graphics to use little r and only months with positive growth

@@
 ---
-title: "Invasion Demography Draft"
+title: "Forecasting Models For Invasive Species (Draft)"
 author: "Gotelli et al."
@@
 Modeling r as a function of environmental variables is an empirical realization of Hutchinson's () definition of the niche: an $n$-dimensional hypervolume describing a combination of variables for which $dN/dt > 0$.
+
+## Case Study #1: Medfly
+
+We begin with the simplest kind of data for demographic modeling: life table analysis of organisms reared im constant laboratory conditions.
+Vargas et al.
+(2000) describe a lab study in which cohorts of the Mediterranean fruit fly Ceritata capitata were raised in one of four different temperature treatments.
+These tephritid flies are native to sub-Saharan Africa, and are serious agricultural pests throughout the world.
+
+In each of the four treatments in the Vargas et al.
+(2000) experiments, cohorts of eggs were reared at at different minimum and maximum temperatures, which were maintained on a daily 9h:9h cycle, with 3 hours of ramping between the maximum and minimum.
+For our analyses, we used the average of the maximum and minimum to characterize the treatment.
+Vargas et al.
+(2000) present many different demographic rates from these experiments.
+We had hoped to model age-specific survivorship and fecundity in each treatment as a function of temperature.
+However, the raw data for these life table analyses were not presented.
+Instead, we have used the average population growth rate r, reported in Table 6 of Vargas et al.
+as the response variable.
+
+Figure 1 shows the relationship between population growth rate and average temperature for each of the 4 treatments.
+The smooth black curve is the best-fit quadratic regression line.
+This model predicts a range of viable temperatures for population growth between 16.7 and 33.2$^{\circ}$C.
+The model predicts maximum population growth rate at a temperature of 24.9$^{\circ}$C, which is close to estimates of an earlier study of maximum net reproductive rate at a constant temperature of 24$^{\circ}$C (Vargas et al.
+1997).
+
+![](../Graphics/Figure1.jpeg)
+
+Next, we used this prediction function to estimate medfly population growth rate across the continental U.S. Because

### 2020-06-28 Scripts/MatrixFunctions.R

@@ -97,7 +97,7 @@ mat_fill <- function(z=1) { # place predicted coefficients in new transition mat
 t2 <- map(t1,mat_fill) %>% # extract population growth rate from 
 map(lambda) %>%
 # t3 <- lapply(t2, function (x) ifelse(x < 0,0,x))
-map(~ifelse(. < 0,0,.))# rescale any values with lambda <0 to 0
+map(~ifelse(. <= 0,0,.))# rescale any values with lambda <0 to 0
 return(t2) 
 
 }
@@ -110,7 +110,8 @@ return(t2)
 lam2_gen <- function(x,y,z){
 . <- est_trans(x,y)
 . <- pred_trans(z,.)
-.[.<0] <- 0  # rescale any values with lambda <0 to 0
+# .[.<0] <- 0  # rescale any values with lambda <0 to 0
+ # .[.<0] <- 0  # rescale any values with lambda <1 to 0
  names(.) <- paste0("V",seq_along(z))
 . <- as.list(.)
 return(.)

### 2020-06-28 Scripts/MattsGraphicsCode.R

@@ -40,9 +40,9 @@ exp_tmp <- c(18.5,24,23.5,29.5)
 # set up temperature input vector for generating continental map
 con_temp <- 32
 
-# for lam2_gen function set up vector of empirical reported lambdas:
+# for lam2_gen function set up vector of empirical reported little_r:
 emp_lam <- c(0.051,0.120,0.137,0.092)
-emp_lam <- exp(emp_lam)
+# emp_lam <- exp(emp_lam)
 ################################################################################
 # download monthly worldclim temp, crop to North America, calc mean temp
 tmin <- getData(name="worldclim",
@@ -70,18 +70,19 @@ lambda_rasts <- mclapply(1:12, function(x, tempRasts){
   # lamb <- lam_gen(rast[keep]) # calc lamba for each temp
   # lamb[lamb<1] <-0 # clip non-increasing values to 0
 #-------------------------------------
-# use lambda2_gen, which is based on lambdas, not transition matrices
+# use lambda2_gen, which is based on little_r, not transition matrices
   lamb <- lam2_gen(x=exp_tmp,
   								 y=emp_lam,
-  								 z=rast[keep]) # calc lamba for each temp
-  lamb[lamb<1] <-0 # clip non-increasing values to 0
+  								 z=rast[keep]) # calc little_r for each temp
+    lamb[lamb<=0] <-NA # clip non-increasing values to 0
 #-------------------------------------
   rast[which(keep)] <- unlist(lamb) # assign lamba back to raster
   return(rast)}, tempRasts=tmean, mc.cores=2)
 lambda_rasts <- do.call(stack, lambda_rasts)
-names(lambda_rasts) <-paste0("medfly", 1:12)
+names(lambda_rasts) <-paste("month", 1:12,sep=" ")
 plot(lambda_rasts)
-
 # calc geometric mean of lambda across the 12 months
-lambdaMap <- calc(lambda_rasts, fun=function(x){prod(x)^(1/length(x))})
-plot(lambdaMap, main="geoMean_of_lambda")
+  # lambdaMap <- calc(lambda_rasts, fun=function(x){prod(exp(x))^(1/length(x))})
+# lambdaMap <- calc(lambda_rasts, fun=function(x){mean(x)})
+lambdaMap <- calc(lambda_rasts, fun=function(x){ifelse(mean(x)>0,mean(x),NA)})
+plot(lambdaMap, main="mean r")

### 2020-06-28 Scripts/NewFigure1Medfly.R

@@ -1,0 +1,81 @@
+# --------------------------------------
+# Updated first graph with simple medfly 
+# growth points
+# 25 Jun 2020
+# NJG
+# --------------------------------------
+library(tidyverse)
+library(logger)
+log_threshold(TRACE)
+mylog <- tempfile()
+log_appender(appender_tee(mylog))
+
+# Simple graph plot for two different 
+# estimation methods for lambda
+source("Scripts/MatrixFunctions.R")
+# set up experimental temperature vector
+exp_tmp <- c(18.5,24,23.5,29.5)
+
+# set up temperature input vector for generating continental map
+con_temp <- 15:35 # temperature range for extrapolation
+
+# set up transition matrices for lam_gen (data from Vargas et al.)
+# fly_18 <- matrix(c(0,0.97,0,0,0,0,0.887,0,0,0,0,1.0,1.35,0,0,0),nrow=4)
+# fly_24 <- matrix(c(0,0.92,0,0,0,0,0.913,0,0,0,0,0.964,3.4,0,0,0),nrow=4)
+# fly_23 <- matrix(c(0,0.92,0,0,0,0,0.962,0,0,0,0,0.870,5.6,0,0,0),nrow=4)
+# fly_29 <- matrix(c(0,0.92,0,0,0,0,0.847,0,0,0,0,0.782,4.0,0,0,0),nrow=4)
+
+# fold into a list
+# fly_dat <- list(fly_18,fly_24,fly_23,fly_29)
+
+# for lam2_gen function set up vector of empirical reported lambdas:
+emp_lam <- c(0.051,0.120,0.137,0.092)
+ # emp_lam <- exp(emp_lam)
+
+# extrapolate lambda from vital rates
+# t1 <- lam_gen(ct=con_temp,
+# 							tm=exp_tmp,
+# 							tr=fly_dat)
+# 	
+# extrapolate lambda from reported lambdas
+t2 <- lam2_gen(x=exp_tmp,
+							 y=emp_lam,
+							 z=con_temp)
+
+# combine into single response vector
+lambda.vec <- unlist(t2)
+names(lambda.vec) <- NULL
+log_info('check this')
+log_trace(cat(lambda.vec),'check here')
+# create temperature scale
+temperature <- rep(con_temp,2)
+
+# create labels for the two series
+# model <- rep(c("Vital Rates Prediction","Lambda Prediction"),each=length(con_temp))
+
+# combine into data frame
[…]

### 2020-06-28 ManuscriptDraft/ManuscriptDraft.Rmd
finish medfly text

@@
 
-Next, we used this prediction function to estimate medfly population growth rate across the continental U.S. Because
+Next, we used this prediction function to estimate medfly population growth rate across the continental U.S.
+For these predictions, we used current WorldClim values for average monthly temperature.
+Figure 2 shows the pixels in which r > 0 for each month of the year, based on the prediction function in Figure 1.
+There is wide spatial variation, with potential medfly distributions covering much of North America in the warmer months, but being absent from the continental U.S.
+during most of the colder months.
+
+![](../Graphics/Figure2.jpeg)
+
+Finally, we assumed for populations to be self-sustaining, they would need to have r > 0 for all 12 months of the year.
+Figure 3 illustrates this pattern, and predicts that the medfly would be restricted to the southern most tip of the Florida peninsula and the Florida Keys.
+
+![](../Graphics/Figure3.jpeg)
+
+A less restrictive model (not shown) would consider viable pixels to be those in which the average population growth rate across all 12 months is greater than 0.
+This model would extend the predicted distribution of the medfly up the Florida peninsula, and down a narrow coastal band of the Gulf Coast.
+However, this model seems unrealistic because the medfly generation time is < 40d and because the cold months occur consecutively in the winter season, so that population growth that occurred during the summer could not sustain local populations through the cold winter months.
+
+How does this forecast from a demographically-based species distribution compare to the actual distribution of the medfly?
+Surprisingly well.
+Medfly populations in the continental U.S.
+are currently found only in southern Florida, which nicely matches our model predictions.
+Medfly has also been eradicated from agricultural areas in southern California, but high levels of watering may have made the microclimate there more comparable to the warm moist habitat of southern Florida.
+
+In contrast, MaxEnt, GARP, and other species distribution models based on occurrence records of the medfly in its native range of sub-Saharan Africa and other introduced sites substantially over-estimate its range in North America.
+These models do not incorporate the acute temperature sensitivity of medfly population growth rate to temperature, and may also be based on occurrences in places that experience less seasonal variability in temperatures than continental North America.
+
+## Case Study #2: *Bromus*

### 2020-06-29 ManuscriptDraft/ManuscriptDraft.Rmd
finish manuscript draft

@@
 date: "6/23/2020"
-output: pdf_document
+output:
+  word_document: default
+  pdf_document: default
 ---
@@
 ## Case Study #2: *Bromus*
+
+As an important alternative to traditional life-table experiments, lambda can be estimated from short-term field measurements of abundance.
+Specifically, if abundance is measured in the same plot in two consecutive time periods, then $log(\frac{N_{t + 1}}{N_{t}}) = r$.
+If abiotic variables such as temperature or precipitation can be measured at the plot scale, these can be used to generate predictive functions as in Equation 2.
+
+We illustrate this method with 7 years of annual census data for cheatgrass....(*need full methods section here from Michael and Alden*).
+The data consist of measurements from experimental snow-fence plots.
+Although some plots were measured for several consecutive years, there are not enough data for a formal time series analysis.
+Therefore, we treated each pair of observations on a single plot in two consecutive years as an independent data point.
+
+For covariates, we analyzed the following abiotic variables: growing season snow, growing season rain, growing season total precipitation, growing season degree-days, winter snow, winter rain, winter total precipitation, winter degree days with no snowpack, winter degree days, snow cover.
+
+We analyzed the relationship between these variables and r using linear and quadratic regression models.
+For most variables, adding a quadratic term substantially improved the r2 and model fit, The quadratic coefficients were usually negative, implying a unimodal peak of population growth at an intermediate level of the variable.
+
+We also included initial population density as a potential covariate.
+If population growth is density-dependent rather than exponential, then growth rate should decline with increasing density.
+In the simplest logistic growth model, the relationship between r and population size is linear.
+We confirmed this linear relationship for the cheatgrass data, with no significant improvement from adding a quadratic term.
+For this logistic growth model, maximum population growth rate is achieved as population size approaches 0.
+The intercept of this model estimates the maximum population growth rate as $r$ = 0.642016 individuals/individual/year, which corresponds to a doubling time of 1.08 years.
+However, many individual plots had recorded estimates much higher than this.
+From this model, we can also estimate the carrying capacity of the population as the population size for which $r$ = 0.
+With a fitted linear slope of -.001326, this estimated $K$ is $0.642016/0.001326 = 484$ individuals per plot.
+
+We selected variables for the final model by evaluating the statistical significance of the individual coefficients, and changes in the adjusted r2, which imposes an AIC penalty on complex models with many parameters.
+We also tried to include interaction terms between the predictor variables, but these never improved the model fit, so they were not retained.
+
+The final best-fitting model included quadratic functions for snow cover and winter rain, and a linear function for initial density (Figure 4):
+
+![](../Graphics/Composite.jpeg)
+
+The final model with these three variables explained 79.1% of the variance in *r*.
+All of the model coefficients were statistically significant, except for the quadratic term for snow cover (p = 0.12).
+
+For purposes of forecasting, we used mapped data layers for snow cover and winter rain.
+We set population size to a constant of 0, so that population growth rate was estimated for an empty landscape in which new populations might become established.
+Although density was set as a constant for the forecasts, it is still important to have it in the model so that the coefficients for snow cover and winter rain are properly estimated.
+
+Figure 5 illustrates a heat map in which r is plotted as a function of snow cover and winter rain.
+This plot is a visualization of the Hutchinsonian niche, because it indicates combinations of abiotic factors for which $dN/dt > 0$ for colonizing populations.
+
+![](../Graphics/HeatMap.jpeg)
+
+Finally, we apply the model using data layers for contemporary snow cover and winter rain.
+The model predictions are in line with the current distribution of Bromus...
+In contrast, more traditional SDMs of Bromus based on species occurrence records suggest that...
+(*need input from Michael and Alden here)
+
+![](../Graphics/bromusLambdaPrediction.jpeg)
+
+These

### 2020-06-29 Scripts/UpdatedBromusRead.R

@@ -1,7 +1,13 @@
+# Latest rendering of figures for best fitting model
+# and final data for Bromus
+# 29 June 2020
+
 # New update of data with Alden's modification for conversion of % snow cover to snow depth
 
 #11 October 2019
-
+library(tidyverse)
+library(patchwork)
+# source("Scripts/BromusLambdaFunction.R")
 # -------------------------------------------------------------------------
 
 # Script for opening and cleaning new Bromus data
@@ -10,7 +16,146 @@
 
 # Read in data ------------------------------------------------------------
 
-z <- read.table("CleanData/Bromus_Clean_V3_Oct-08-19.csv",sep=",",
+df <- read.table("CleanData/Bromus_Clean_V3_Oct-08-19.csv",sep=",",
                 stringsAsFactors=FALSE, header=TRUE)
-str(z)
+str(df)
 
+df <- df[,-ncol(df)] # remove empty final column 
+df_clean <- df[complete.cases(df),] # remove 4 missing lambda
+df_preds <- df_clean[,8:18] # create dataframe of predictor variables 
+df_clean$r <- log(df_clean$Lambda) # create response variable r from log(lambda)
+
+best_preds <- df_preds[,c(6,7,11,8,3,1)]
+y <- df_clean$r
+x_1 <- best_preds[,1] 
+x_2 <- best_preds[,2] 
+x_3 <- best_preds[,3] 
+x_4 <- best_preds[,4] 
+x_5 <- best_preds[,5] 
+x_6 <- best_preds[,6]
+
+final_df <- data.frame(r=y, snow_cover=x_1,
+                       win_precip_snow=x_2,
+                       win_precip_tot=x_3,
+                       win_precip_rain=x_4,
+                       grow_seas_precip_rain=x_5,
+                       density=x_6)
+
+# best fitting final model
+model_reduced <- lm(data=final_df,
+r~poly(snow_cover,2, raw=TRUE) + 
+poly(win_precip_rain,2, raw=TRUE) + 
+poly(density,1, raw=TRUE))
+print(model_reduced)
+
+
+print(summary(model_reduced))
+
+#-------------------------------------
+# FUNCTION pred_brom_lam
+# predict lambda based on snow cover, precip, and density
+# input: f=fitted model, new= data frame for prediction
[…]

### 2020-07-07 Scripts/HouseSparrowAnalyses.Rmd
add sparrow data and initial analysis

---
title: "Sparrow initial analyses"
author: "Nick Gotelli"
date: "7/7/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
This analysis uses data from Jesse Krause on house sparrow growth rates as a function of precipitation and temperature, compiled from two different studies im Europe. The original compiled data are contained in `House Sparrow Lambda.xlsx`.

### Build Data Frame
```{r}
lambda <- c(0.8,0.78,0.78,2.6,1.75,0.8,1.03,1.2,0.8,0.95,0.78,0.7,0.99,1.1,1.4,1.1,0.8,1.1,0.9,0.97,0.9,0.58,0.82)
r = log(lambda) # convert to little r

temp <- c(11,9,10,10.5,10.8,9.9,9.5,12,10,10,9.8,10,20.8,20.2,21.5,22.3,19.7,20.2,20.4,21.7,21,21.1,20)
precip <- c(1,2,3,1,2,2,1,4,4,1,4,4,0.816666667,0.583333333,1.366666667,0.95,1.423333333,1.666666667,2.3,1.82,1.59,3.733333333,2.756666667)

site <- rep(c("Norway","Hungary"),each=12)
site <- site[-24] #trim last element

df <- data.frame(r,temp,precip,site)
```
### Set up Model and Graph Functions
```{r}
# Function quad_fit ------------------------------------------------------
# inputs: response variable, predictor variable
# outputs: adjusted r2 for linear and quadratic regression models

quad_fit <- function(x,y){
  M1 <- summary(lm(y~x))
  M2 <- summary(lm(y~poly(x,2,raw=TRUE)))
  return(list(M1,M2))
}
# -------------------------------------------------------------------------
# Function quad Graph
# Function quad_plot ------------------------------------------------------
# inputs: response variable, predictor variable
# outputs: ggplot graph with data and fitted line

quad_plot <- function(x,y,z){
  df <- data.frame(x,y,z)
  ggplot(data=df,aes(x=x,y=y)) +
  geom_point(shape=z,size=4) +
  stat_smooth(method="lm", se=TRUE, fill=NA,
                formula=y ~ poly(x, 2, raw=TRUE),colour="red") +
    stat_smooth(method="lm", se=TRUE, fill=NA,
                formula=y ~ poly(x, 1, raw=TRUE),colour="blue") +
    ylab("r") +
    xlab("Environmental Predictor") +
    theme_bw(base_size = 20)
}
```

### Temperature Analysis

```{r}
quad_fit(x=df$temp,y=df$r)
quad_plot(x=df$temp,y=df$r,z=df$site)
```

Although the fitted curve does have an optimal intermediate temperature, neither the linear nor the quadratic model explain any of the variance in r. Statistically, there is no justification for including temperature as a predictor in the final model. However, we could still use it in the model because it does predict a range of viable temperatures between 10.5 and 20.3.

### Precipitation Analysis

```{r}
quad_fit(x=df$precip,y=df$r)
quad_plot(x=df$precip,y=df$r,z=df$site)
```

For the precipitation data, the best fitting model is a linear model in which population growth rate declines with increasing rainfall. The quadratic term is non-significant and the coefficient is positive, which means the curve does not have an internal peak. Using the linear model, we get a simple prediction of positive population growth rate when precipitation is less than `r 0.2206/0.1201` mm/day.

Does that seem biologically realistic?

For both variables, an additional question is how we translate these units into measurements at the continental scale. Do we want yearly averages or just averages during the breeding season? For the medfly, we fitted the model to monthly averages because the generation time is so short, then averaged growth rates over the entire year. For *Bromus*, I believe we were using annual measures of precipitation to scale with annual measures of change in population size.

### 2024-02-22 Scripts/SupplementB.Rmd
created Supplement B for submission

--- 2019-09-15 Scripts/BromusRollingUpdateFile.Rmd
@@
 ---
-title: '*Bromus* Density Dependence Analysis'
-author: "Nick Gotelli"
-date: "4 September 2019"
+title: 'Supplement B'
+author: "Bromus predictor variable analysis"
+date: "Gotelli et al. 22 Feb 2024"
 output:
@@
 ```
-
-## 15 September 2019
-This update adds raw to all poly statements, tries adding an interaction term for snow and precip, and constructs heat maps for visualizing the Hutchinsonian niche.
-
-## Update status
-This update uses the expanded data set from Alden and Catherine and cleans up the coefficients in the quadratic function
-
-## Data sources and cleaning
-These data were provided by Alden Griffith and Michael Loic.
-After talking with Alden, we decided to use $\lambda$ values that had been estimated from changes in the number of individuals in marked plots, rather than analysis of stage transition matrices.
-For a number of control plots, Alden and Michael have estimated lambda $(N_{t + 1}/N_{t})$ based on annual counts of cheatgrass for two consecutive years.
-For some of the plots, there are actually 7 consecutive years of data, but I am ignoring plot identity, and simply using $\lambda$ calculated for pair of consecutive abundances.
-For each of these measures, the variable `Starting.Density` gives $N_t$, the abundance in the first year.
-
-For a simple logistic growth equation, there should be a negative linear relationship between  $r =        log(\lambda)$ and $N_t$.
-
-I cleaned the initial data file (`Bromus Lambda and Climate Data.csv`), moving a notes column to initial comment lines, and adding 4 missing values as `NA`.
-The modified data are in `CleanBromusData.csv`.
 
@@
 ## Initial regression models
-With the data frame set up, the first analysis was to use each of the predictor variables (snow, degree-days, precipitation, etc.), with $r$ as the response variable.
-For each variable, I built a linear and a quadratic regression model.
 
@@
 ## First screening of predictor variables
-In this table, each row is a different model with a single predictor variable, and the data frame shows the adjusted r^2^ for each model.
+Each row is a different model with a single predictor variable, and the data frame shows the adjusted r^2^ for each model.
 The two columns are for a linear model $(y \sim x)$ and a quadratic model $(y \sim x + x^2)$.
@@
 
-I had originally intended to include one of the temperature variables, but they perform poorly compared to precip and snow cover variables, so we will start with just these initial 4.
-Many of these snow variables will be intercorrelated, so we should be able to shrink down to a smaller set of predictors.
-These results are qualitatively similar to what we found with the initial data set, but the density dependence is weaker in this expanded data set.
-
 ## Simple linear density dependence model
-
-We start by ignoring all the climate variables and just fit the collection of annual $\lambda$ values to the linear function we expect for density dependence in a plain-vanilla logistic growth model:
 
@@
 ```
-
-These results look pretty nice.
-The relationship is roughly linear, and we pick up a sizeable chunk of the variation in $r$.
-The intercept is 0.642016, which corresponds to the maximum $r$ that occurs with no crowding ($N_t$ = 0).
-This is the constant $r$ that appears in the logistic growth equation.
-It is large ($\lambda = exp(r) = 1.9$), but reasonable for an annual plant.
-You can see there are some low-density plots that have even higher values than this.
-
-The other parameter of interest is to set $r = 0$ and solve for $N$. This is the estimate of the carrying capacity $K$ in the logistic growth equation. For these data, we get K = 484 individuals.
-
-Alden and Michael, do these values of $r = 1.64 individuals/individual\times year$ and $K = 484 individuals$ seem biologically reasonable for *Bromus**?
 
@@
 
-The quadratic models all look good, except this one for growing season rain, which has a positive quadratic term.
-
 ## Saturated regression model with all predictors
@@
 
-Based on these initial results, we can winnow out a few variables to start with:
-
 ```{r}
@@
 
-For this model, we will try dropping out `grow_seas_precip_rain` simply because that variable has a positive quadratic term, which is not biologically realistic, and it only adds about 2% variance explained.
+```{r}
+model_reduced2 <- lm(data=final_df,r~poly(snow_cover,2, raw=TRUE) +
+                                              poly(win_precip_rain,2, raw=TRUE) +
+                                              poly(density,1, raw=TRUE))
 
-```{r}
-model_reduced2 <- summary(lm(data=final_df,r~poly(snow_cover,2, raw=TRUE) +
-                                              poly(win_precip_rain,2, raw=TRUE) +
-                                              poly(density,1, raw=TRUE)))
-
-print(model_reduced2)
+print(summary(model_reduced2))
 ```
@@
 This reduced model has only 3 variables: snow cover, winter rain, and initial density, and accounts for almost 80% of the variation in population growth rate.
-That's about as good as it gets for regression models with observational data.
-
-The next step will be to build a prediction function for Matt, who will then need to set up data layers for snow cover and winter rain.
-In the prediction model, we will set density to 0 because we are looking at colonization potential of *Bromus* in an empty continental landscape.
 
 ## Reduced model with interaction term
-
-For this model, we will try an interaction between the linear components of snow cover and rain
 
@@
 
-Since the interaction term doesn't add anything, we will not include it and stick with the simple reduced model.
+## Prediction function
+
+```{r}
+#-------------------------------------
+# FUNCTION pred_brom_lam
+# predict lambda based on snow cover, precip, and density
+# input: f=fitted model, new= data frame for prediction
+# output: new model coefficient
+pred_brom_lam <- function(new=NULL,f=model_reduced2){
+            if (is.null(new)){
+            	x1 <- runif(10,0,1)
+            	x2 <- runif(10,0,200)
+            	x3 <- rep(0,10)
+            	new<- data.frame(snow_cover=x1,
+            	                win_precip_rain=x2,
+            	                density=x3)}
+
+	z <- predict(object=f, newdata=new)
+	return(z)
+
+}
+#-------------------------------------
+```
 
 ## Heat map for visualizing lambda values
+
+```{r}
+# construct data frame for sweeping parameter space of snow cover and winter precip
+# density is held constant at N = 0
+# snow cover ranges [0,1]
+# winter precip rain ranges [0,200]
+vec_len <- 30
+snow_vec <- seq(0,1,length.out=vec_len)
+precip_vec <- seq(0,200,length.out=vec_len)
+hutchinson_df <- expand.grid(snow_cover=snow_vec,win_precip_rain=precip_vec)
+hutchinson_df$density <- 0
+
+# this data frame has the structure needed as input for creating a continental map
+str(hutchinson_df)
+
+# now run the pred_brom_lam() function to predict r and bind to input data frame
+heat_map_data <- cbind(hutchinson_df,r=pred_brom_lam(new=hutchinson_df))
+str(heat_map_data)
+
+# construct heat map
+heat_map <- ggplot(data=heat_map_data,mapping=aes(x=snow_cover,y=win_precip_rain,fill=r))
+heat_map + geom_tile() + scale_fill_viridis_c()
+```

### 2024-11-07 Scripts/maxentModeling.R
Initial commit

@@ -1,0 +1,43 @@
+library(dismo)
+library(mapdata)
+
+# medfly
+medfly <- gbif(genus="Ceratitis", 
+               species="capitata", 
+               geo=T, 
+               removeZeros=T)#,
+#ext=extent(c(-180, -30, 0, 90)))
+
+#medfly <- subset(medfly, country %in% c("United States", "Canada", "Mexico"))
+
+# plot
+map('worldHires')#,  c('USA', 'Canada', 'Mexico'), xlim=c(-180,-30))
+points(medfly$lon, medfly$lat, pch=20, col=rgb(0,0,1,0.5))
+
+
+# cheat grass
+cheatgrass <- gbif(genus="Bromus", 
+                   species="tectorum", 
+                   geo=T,
+                   removeZeros=T)#,
+#ext=extent(c(-180, -30, 0, 90)))
+
+# plot
+map('worldHires')#,  c('USA', 'Canada', 'Mexico'), xlim=c(-180,-30))
+points(cheatgrass$lon, cheatgrass$lat, pch=20, col=rgb(0,0,1,0.5))
+
+
+
+# house sparrow
+housesparrow <- gbif(genus="Passer", 
+                     species="domesticus", 
+                     geo=T,
+                     removeZeros=T,
+                     args=c("year"==2021))#,
+                     #ext=extent(c(-70, -50, 30, 40)))
+
+#housesparrow <- subset(housesparrow, country %in% c("United States", "Canada", "Mexico"))
+
+# plot
+map('worldHires')#,  c('USA', 'Canada', 'Mexico'), xlim=c(-180,-30))
+points(housesparrow$lon, housesparrow$lat, pch=20, col=rgb(0,0,1,0.5))

### 2024-11-07 Scripts/gbifMaps.R
minor edit to house sparrow

@@ -35,7 +35,8 @@ housesparrow <- gbif(genus="Passer",
                    species="domesticus", 
                    geo=T,
                    removeZeros=T,
-                   ext=extent(c(-180, -30, 0, 90)))
+                   args=c("year"==2021),
+                   ext=extent(c(-70, -50, 30, 40)))
 
 housesparrow <- subset(housesparrow, country %in% c("United States", "Canada", "Mexico"))
 

### 2024-11-08 Scripts/BromusRollingUpdateFile.Rmd
Start building final functions
Initial commit

@@
 ```{r}
-model_reduced2 <- summary(lm(data=final_df,r~poly(snow_cover,2, raw=TRUE) +
+model_reduced2 <- lm(data=final_df,r~poly(snow_cover,2, raw=TRUE) +
                                               poly(win_precip_rain,2, raw=TRUE) +
-                                              poly(density,1, raw=TRUE)))
+                                              poly(density,1, raw=TRUE))
 
-print(model_reduced2)
+print(summary(model_reduced2))
 ```
@@
 
+## Prediction function
+
+```{r}
+#-------------------------------------
+# FUNCTION pred_brom_lam
+# predict lambda based on snow cover, precip, and density
+# input: f=fitted model, new= data frame for prediction
+# output: new model coefficient
+pred_brom_lam <- function(new=NULL,f=model_reduced2){
+            if (is.null(new)){
+            	x1 <- runif(10,0,1)
+            	x2 <- runif(10,0,200)
+            	x3 <- rep(0,10)
+            	new<- data.frame(snow_cover=x1,
+            	                win_precip_rain=x2,
+            	                density=x3)}
+
+	z <- predict(object=f, newdata=new)
+	return(z)
+
+}
+#-------------------------------------
+```
+
 ## Heat map for visualizing lambda values
+
+```{r}
+# construct data frame for sweeping parameter space of snow cover and winter precip
+# density is held constant at N = 0
+# snow cover ranges [0,1]
+# winter precip rain ranges [0,200]
+vec_len <- 30
+snow_vec <- seq(0,1,length.out=vec_len)
+precip_vec <- seq(0,200,length.out=vec_len)
+hutchinson_df <- expand.grid(snow_cover=snow_vec,win_precip_rain=precip_vec)
+hutchinson_df$density <- 0
+
+# this data frame has the structure needed as input for creating a continental map
+str(hutchinson_df)
+
+# now run the pred_brom_lam() function to predict r and bind to input data frame
+heat_map_data <- cbind(hutchinson_df,r=pred_brom_lam(new=hutchinson_df))
+str(heat_map_data)
+
+# construct heat map
+heat_map <- ggplot(data=heat_map_data,mapping=aes(x=snow_cover,y=win_precip_rain,fill=r))
+heat_map + geom_tile() + scale_fill_viridis_c()
+```

### 2024-11-08 Scripts/BromusUpdate06Augl2019.Rmd
Initial commit

--- 2019-09-04 Scripts/BromusUpdate0Sep2019.Rmd
@@
 author: "Nick Gotelli"
-date: "4 September 2019"
+date: "7/30/2019"
 output:
@@
 ## Updatre status
-This update uses the expanded data set from Alden and Catherine and cleans up the coefficients in the quadratic function
+This update ignores the puzzling quadratic coefficient for density dependence, and just moves on to fitting the other abiotic predictors and creating a heat map of the parameter space.
 
@@
 # Get input data ----------------------------------------------------------
-df <- read.table(file="../CleanData/Bromus_Clean_V2_Aug-28-19.csv",
-                 sep=",",header=TRUE,stringsAsFactors=FALSE)
+df <- read.csv(file="../CleanData/CleanBromusData.csv",comment.char="#")
 df <- df[,-ncol(df)]                 # remove empty final column
@@
   M1 <- summary(lm(y~x))
-  M2 <- summary(lm(y~poly(x,2,raw=TRUE)))
+  M2 <- summary(lm(y~poly(x,2)))
   return(c(M1$adj.r.squared,M2$adj.r.squared))
@@
 
-The quadratic model often fits better than the corresponding linear model (even after imposing a penalty for the additional parameter), so we have good theoretical grounds and empirical support for going with the quadratic models.
-However, the quadratic model does not provide much improvement for the density dependence, so we will go with the linear model, which corresponds to a plain-vanilla logistic growth equation.
+A couple things jump out from these univariate regressions.
+First, the quadratic model often fits better than the corresponding linear model (even after imposing a penalty for the additional parameter), so we have good theoretical grounds and empirical support for going with the quadratic models.
 
-The strongest predictors were:
+The four strongest predictors were:
 
-- snow cover (quadratic r^2^ = 0.759)
-- winter precip snow (quadratic r^2^ = 0.612)
-- winter precip total (quadratic r^2^ = 0.555)
-- winter precip rain (quadratic r^2^ = 0.509)
-- growing season precip rain (quadratic r^2^ = 0.469)
-- initial density (linear r^2^ = 0.368)
+- snow cover (quadratic r^2^ = 0.838)
+- winter precip total (quadratic r^2^ = 0.661)
+- initial density (linear r^2^ = 0.638)
+- growing season precipitation rain (quadratic r^2^ = 0.559)
 
 I had originally intended to include one of the temperature variables, but they perform poorly compared to precip and snow cover variables, so we will start with just these initial 4.
-Many of these snow variables will be intercorrelated, so we should be able to shrink down to a smaller set of predictors.
-These results are qualitatively similar to what we found with the initial data set, but the density dependence is weaker in this expanded data set.
 
@@
 # snow.cover (6)
-# winter precip snow (7)
-# winter precip total (11)
-# winter precip rain (8)
-# growing season precip rain (3)
 # density (1)
-best_preds <- df_preds[,c(6,7,11,8,3,1)]
+# winter precip total. (11) 
+# grow season precip rain (3)
+
+best_preds <- df_preds[,c(6,1,11,3)]
 
@@
 x_4 <- best_preds[,4]
-x_5 <- best_preds[,5]
-x_6 <- best_preds[,6]
 
@@
                        snow_cover=x_1,
-                       win_precip_snow=x_2,
-                       win_precip_tot=x_3,
-                       win_precip_rain=x_4,
-                       grow_seas_precip_rain=x_5,
-                       density=x_6)
+                       density=x_2,
+                       win_precip <- x_3,
+                       gs_rain <- x_4)
 
@@
 The relationship is roughly linear, and we pick up a sizeable chunk of the variation in $r$.
-The intercept is 0.642016, which corresponds to the maximum $r$ that occurs with no crowding ($N_t$ = 0).
+The intercept is 1.206162, which corresponds to the maximum $r$ that occurs with no crowding ($N_t$ = 0).
 This is the constant $r$ that appears in the logistic growth equation.
-It is large ($\lambda = exp(r) = 1.9$), but reasonable for an annual plant.
+It is large ($\lambda = exp(r) = 3.34$), but reasonable for an annual plant.
 You can see there are some low-density plots that have even higher values than this.
 
-The other parameter of interest is to set $r = 0$ and solve for $N$. This is the estimate of the carrying capacity $K$ in the logistic growth equation. For these data, we get K = 484 individuals.
+The other parameter of interest is to set $r = 0$ and solve for $N$. This is the estimate of the carrying capacity $K$ in the logistic growth equation. For these data, we get K = 714 individuals.
 
-Alden and Michael, do these values of $r = 1.64 individuals/individual\times year$ and $K = 484 individuals$ seem biologically reasonable for *Bromus**?
+Alden and Michael, do these values of $r = 1.21 individuals/individual\times year$ and $K = 714 individuals$ seem biologically reasonable for *Bromus**?
 
@@
 
-## Quadratic model for winter precipitation snow
+As with the effects of initial density, the intercept for the quadratic model still does not look correct to me, suggesting that either I am doing something wrong in the code or I don't understand the output from a linear model.
+
+## Quadratic model for winter precipitation
 
 ```{r}
-model_win_precip <- summary(lm(data=final_df,r~poly(win_precip_snow,2)))
+model_win_precip <- summary(lm(data=final_df,r~poly(win_precip,2)))
 print(model_win_precip)
-ggplot(data=final_df,aes(x=win_precip_snow,y=r)) +
+ggplot(data=final_df,aes(x=win_precip,y=r)) +
   geom_point() +
@@
 
-## Quadratic model for winter precip total
+This also looks reasonable (except for the intercept), but it will be highly correlated with snow cover, and it is not clear what will happen with both variables in the model.
+
+## Quadratic model for growing season rainfall
 
 ```{r}
-model_win_precip_tot <- summary(lm(data=final_df,r~poly(win_precip_tot,2)))
-print(model_win_precip_tot)
-ggplot(data=final_df,aes(x=win_precip_tot,y=r)) +
+model_gs_rain <- summary(lm(data=final_df,r~poly(gs_rain,2)))
+print(model_gs_rain)
+ggplot(data=final_df,aes(x=gs_rain,y=r)) +
   geom_point() +
@@
 
-## Quadratic model for winter precip rain
-
-```{r}
-model_win_precip_rain <- summary(lm(data=final_df,r~poly(win_precip_rain,2)))
-print(model_win_precip_rain)
-ggplot(data=final_df,aes(x=win_precip_rain,y=r)) +
-  geom_point() +
-  stat_smooth(method="lm", se=TRUE, fill=NA,
-                formula=y ~ poly(x, 2, raw=TRUE),colour="red") 
-```
-
-## Quadratic model for growing season precip rain
-
-```{r}
-model_grow_seas_precip_rain <- summary(lm(data=final_df,r~poly(grow_seas_precip_rain,2)))
-print(model_grow_seas_precip_rain)
-ggplot(data=final_df,aes(x=grow_seas_precip_rain,y=r)) +
-  geom_point() +
-  stat_smooth(method="lm", se=TRUE, fill=NA,
-                formula=y ~ poly(x, 2, raw=TRUE),colour="red") 
-```
-
-The quadratic models all look good, except this one for growing season rain, which has a positive quadratic term.
+It is starting to look like the intercept for these polynomial models is always -0.05.
+I will let that go for now, but we will need to get this straightened out.
+The growing season rainfall has an ugly inverted quadratic (caused by the positive coefficient for the x^2^ term).
+Let's see how it performs in a multiple regression model.
 
@@
 ```{r}
-model_saturated <- summary(lm(data=final_df,r~poly(snow_cover,2, raw=TRUE) +
-                                              poly(win_precip_snow,2, raw=TRUE) +
-                                              poly(win_precip_tot,2, raw=TRUE) +
-                                              poly(win_precip_rain,2, raw=TRUE) +
-                                              poly(grow_seas_precip_rain,2, raw=TRUE) +
-                                              poly(density,1)))
+model_saturated <- summary(lm(data=final_df,r~poly(gs_rain,2) +
+                                              poly(density,1) +
+                                              poly(snow_cover,2) +
+                                              poly(win_precip,2)))
 
@@
 ```
-
-## Reduced model
-
-Based on these initial results, we can winnow out a few variables to start with:
-
-```{r}
-model_reduced <- summary(lm(data=final_df,r~poly(snow_cover,2, raw=TRUE) +
-                                              poly(win_precip_snow,2, raw=TRUE) +
-                                              poly(win_precip_tot,2, raw=TRUE) +
-                                              poly(density,1)))
-
-print(model_reduced)
-```
-
-## Reduced model #2
-
-For this model, we will try dropping out `grow_seas_precip_rain` simply because that variable has a positive quadratic term, which is not biologically realistic, and it only adds about 2% variance explained.
-
-```{r}
-model_reduced2 <- summary(lm(data=final_df,r~poly(snow_cover,2, raw=TRUE) +
-                                              poly(win_precip_rain,2, raw=TRUE) +
-                                              poly(density,1)))
-
-print(model_reduced2)
-```
-
-This reduced model has only 3 variables: snow cover, winter rain, and initial density, and accounts for almost 80% of the variation in population growth rate.
-That's about as good as it gets for regression models with observational data.
-
-The next step will be to build a prediction function for Matt, who will then need to set up data layers for snow cover and winter rain.
-In the prediction model, we will set density to 0 because we are looking at colonization potential of *Bromus* in an empty continental landscape.

### 2024-11-08 Scripts/BromusUpdate30Jul2019.Rmd
Initial commit

@@
 ---
-title: '*Bromus* Initial Analysis'
+title: '*Bromus* Density Dependence Analysis'
 author: "Nick Gotelli"
@@
 These data were provided by Alden Griffith and Michael Loic.
-After talking with Alden, we decided to use lambda values that had been estimated from changes in the number of individuals in marked plots, rather than analysis of stage transition matrices.
-For a number of control and snow-manipulated plots, we have the estimated lambda $(N_{t + 1}/N_{t})$ based on annual counts of cheatgrass for 7 consecutive years.
-For incorporating density dependence, we will use $log(\lambda) = r$.
+After talking with Alden, we decided to use $\lambda$ values that had been estimated from changes in the number of individuals in marked plots, rather than analysis of stage transition matrices.
+For a number of control plots, Alden and Michael have estimated lambda $(N_{t + 1}/N_{t})$ based on annual counts of cheatgrass for two consecutive years.
+For some of the plots, there are actually 7 consecutive years of data, but I am ignoring plot identity, and simply using $\lambda$ calculated for pair of consecutive abundances.
+For each of these measures, the variable `Starting.Density` gives $N_t$, the abundance in the first year.
+
+For a simple logistic growth equation, there should be a negative linear relationship between  $r =        log(\lambda)$ and $N_t$.
 
@@
 ## Initial regression models
-With the data frame set up, the first analysis was to use each of the predictor variables (snow, degree-days, precipitation, etc.), with $\lambda$ as the response variable.
+With the data frame set up, the first analysis was to use each of the predictor variables (snow, degree-days, precipitation, etc.), with $r$ as the response variable.
 For each variable, I built a linear and a quadratic regression model.
@@
 - snow cover (quadratic r^2^ = 0.838)
+- winter precip total (quadratic r^2^ = 0.661)
 - initial density (linear r^2^ = 0.638)
-- winter precip total (quadratic r^2^ = 0.661)
 - growing season precipitation rain (quadratic r^2^ = 0.559)
@@
 I had originally intended to include one of the temperature variables, but they perform poorly compared to precip and snow cover variables, so we will start with just these initial 4.
-The linear regression of r versus initial density is the relationship for a simple logistic growth model.
-The performance of the quadratic model was virtually identical, so there is no evidence for Allee effects or other more complex functions relating per capita growth rate to density.
-So we will stick with the linear model.
 
-I wanted to have predictors related to snow, temperature, and precipitation, as those seem like the best suspects for what is driving distribution.
-Based on the adjusted r^2^, the best 3 predictors were `Snow.Cover`, `Grow.Season.Precip.Total`, and `Grow.Season.DD.all.days`.
-There was also a significant effect of `Starting.Density`, which I will discuss near the end.
+## Simple linear density dependence model
 
-## Initial multiple regression model
-
-Now taking those 3 best predictors based on snow, precip, and temperature, here is the full model:
+We start by ignoring all the climate variables and just fit the collection of annual $\lambda$ values to the linear function we expect for density dependence in a plain-vanilla logistic growth model:
 
 ```{r}
-str(df_preds)
 # best variables are:
@@
 print(model_dd)
+ggplot(data=final_df,aes(x=density,y=r)) +
+  geom_point() +
+  stat_smooth(method="lm", se=TRUE, fill=NA,
+                formula=y ~ poly(x, 1, raw=TRUE),colour="red") 
+```
+
+These results look pretty nice.
+The relationship is roughly linear, and we pick up a sizeable chunk of the variation in $r$.
+The intercept is 1.206162, which corresponds to the maximum $r$ that occurs with no crowding ($N_t$ = 0).
+This is the constant $r$ that appears in the logistic growth equation.
+It is large ($\lambda = exp(r) = 3.34$), but reasonable for an annual plant.
+You can see there are some low-density plots that have even higher values than this.
+
+The other parameter of interest is to set $r = 0$ and solve for $N$. This is the estimate of the carrying capacity $K$ in the logistic growth equation. For these data, we get K = 714 individuals.
+
+Alden and Michael, do these values of $r = 1.21 individuals/individual\times year$ and $K = 714 individuals$ seem biologically reasonable for *Bromus**?
+
+## Quadratic density dependence model
+
+The linear equation we fit above looks pretty good, but I wanted to check and see if there is any improvement by adding a quadratic term.
+The quadratic term might pop up if we have Allee effects generating inverse density dependence at low densities.
+This can occur in social animals with behaviors such as cooperative predator defense, cooperative hunting, and mate searching.
+For plants, you could see if it clumps of plants act as pollinator magnets and attract proportionally more pollinators than solitary plants.
+That doesn't seem likely for *Bromus*, but it is good to check the model.
+
+```{r}
 model_dd <- summary(lm(data=final_df,r~poly(density,2)))
@@
                 formula=y ~ poly(x, 1, raw=TRUE),colour="blue")
-
-model_1 <- summary(lm(y~poly(x_1,2) + x_2 + poly(x_3,2) + poly(x_4,2)))
-print(model_1)
-
 ```
 
-## 2-variable model with an interaction term
+With one exception (see below), this output looks reasonable.
+The quadratic term is marginally non-significant (*p* = 0.10, and the quadratic fit does not look very different from the linear fit, except at very high densities, where the density dependence looks a little weaker than predicted by the linear model.
+Based on this comparison, I am comfortable with the linear model (a.k.a.
+logisitic growth) for estimating parameters for snow cover and other abiotic variables.
+Following up on a suggestion from Alden, in the continental forecasting model, we will actually fix N=0 for all plots.
+That way we are predicting lambda as a function of environmental variables for *Bromus* that are colonizing an empty pixel.
+There will not be any density effect in the continental forecast, but we do want the density effect when we fit the prediction functions to make sure we are not confounding abiotic effects with density effects in the original *Bromus* data.
 
-Although the precipitation variable was initially a strong predictor, its effect is non-significant with both snow cover and degree days in the model, so I dropped it at this point.
-The next step was to try a 2-variable model but include an interaction term between snow cover and degree days:
+## Trouble in the quadratic model
 
-```{r}
-# Try adding an interaction term -----------------------------------------
+And now for the fly in the ointment.
+Take a look at the intercept coefficient in the fitted quadratic model.
+Its value is -0.0500.
+That seems dead wrong to me.
+Why isn't it something slightly larger than 1.2, the intercept for the linear model?
+In the graph you can see that those two estimates should be very similar to one another for the point at which N = 0.
 
-model_3 <- summary(lm(y~poly(x_1,2) + poly(x_3,2) + x_1*x_3))
-print(model_3)
-```
+I can't see why the intercept estimates should be so far off. Note that this is a simple least-squares solution, so I am not using any kind of iterative fitting of a non-linear equation.
 
-## Final 2-variable prediction model
-
-However, the interaction term does not add anything. So, we are left with just a two-variable quadratic model for predicting $\lambda$:
-
-```{r}
-model_2 <- summary(lm(y~poly(x_1,2) + poly(x_3,2)))
-print(model_2)
-
-```
-In this model, all of the coefficients are significantly different from zero, and the model explains over 50\% of the variance in $\lambda$, which is decent.
-For now, that is the model we will start with, although it will be easy enough to change if there are other variables you want to consider.
-
-## Functions for fitting and predicting with new values
-
-The next two functions are used to fit the model to the data, and then to predict new values of $\lambda$ based on new inputs of snow cover or degree days.
-
-```{r}
-# Function Lambda Model -----------------------------------------------
-# inputs: lambda vector, and data frame of predictor variables
-# outputs: linear model for forecasting
-
-lambda_model <- function(y=df_clean$r,z=best_preds){
-  y <- df_clean$r
-  x_1 <- z[,1] # Grow.Season.DD.all.days
-  x_2 <- z[,3]  # Snow.Cover
-  forecast_model <- lm(y~poly(x_1,2) + poly(x_2,2))
-}
-
-# Function Lambda predict -------------------------------------------------
-# inputs: linear model, degree days (vector), snow cover (vector)
-# outputs: predicted lambda
-
-lambda_predict <- function(m=my_model,degree_days=c(800,600),snow=c(0.83,.74)) {
-
-z <- predict(object=m, newdata=data.frame(x_1=degree_days,x_2=snow))
-z[z<0] <- 0   # reset to 0 any predicted negative values for lambda
-return(z)
-}
-
-```
-
-## Using the functions for forecasting continental distributions
-
-These two functions will make it possible to generate predictions for any values of snow cover and degree days.
-For example, here the functions are run for two arbitrary vectors of 10 elements each snow cover and degree days.
-The output is the estimated $\lambda$ for each pair of the variables:
-
-```{r}
-lambda_predict(m=lambda_model(),degree_days=sample(600:800,10),snow=runif(10))
-```
-
-When we get to the next step, Matt will use this line of code to generate an estimated lambda for any map location for which we have the snow and degree days measurements.
-
-## Sweeping the parameters and constructing a heat map for $\lambda$
-
-Before we build the continental map, it is instructive to sweep the parameters and visualize the response surface of $\lambda$.
-To set limits on snow cover and degree days, I will consider snow cover between 0 and 100%, and will expand the degree day measurements by 50% beyond the measured limits in Alden and Michael's experiment:
-
-```{r}
-# set limits for parameter inputs
-snow_low <- 0.0
-snow_high <- 1.0
-dd_low <- 0.5*min(x_1)
-dd_high <- 1.5*max(x_1)
-
-# create parameter data frame with column for output
-param_df <- expand.grid(snow_cover=seq(from=snow_low, to=snow_high, length.out=20),
-                        degree_days=seq(from=dd_low, to=dd_high,length.out=20))
-param_df$r <- NA
-
-# apply function to each row
-for (i in 1:nrow(param_df)){
-  param_df[i,3] <- lambda_predict(m=lambda_model(),
-                                  degree_days=param_df[i,2],
-                                  snow=param_df[i,1])
-}
-
-```
-
-## Problems with the model
-Scanning through the completed `param_df` data frame reveals impossibly high values of lambda for most parameter combinations. The problem is coming from the degree days data:
-
-```{r}
-qplot(x=best_preds[,1],y=df_clean$r)
-
-```
-
-Because of the cluster of small values at 800 degree days, the quadratic term in the regression model has a positive coefficient, so $\lambda$ increases at both high and low degree days.
-
-I don't see an easy solution, so for now, I am going to drop back to a model just using snow cover, which has a negative coefficient for the quadratic term and therefore decreases lambda at more extreme values.
-
-## Simple snow cover model
-So, here are the reworked functions just using snow cover:
-```{r}
-# Function Lambda Model2-----------------------------------------------
-# inputs: lambda vector, and data frame of predictor variables
-# outputs: linear model for forecasting
-
-lambda_model2 <- function(y=df_clean$r,z=best_preds){
-  y <- df_clean$r
-  # x_1 <- z[,1] # Grow.Season.DD.all.days
-  x_2 <- z[,3]  # Snow.Cover
-  forecast_model <- lm(y~ poly(x_2,2))
-}
-
-# Function Lambda predict2 -------------------------------------------------
-# inputs: linear model, degree days (vector), snow cover (vector)
-# outputs: predicted lambda
-
-lambda_predict2 <- function(m=my_model,snow=c(0.83,.74)) {
-
-z <- predict(object=m, newdata=data.frame(x_2=snow))
-z[z<0] <- 0   # reset to 0 any predicted negative values for lambda
-return(z)
-}
-
-param_df2 <- data.frame(snow_cover=seq(from=snow_low, to=snow_high, length.out=20),
-                        lambda=NA)
-```
-
-Now with these simplified functions, we run the model once again:
-
-```{r}
-
-# apply function to each row
-for (i in 1:nrow(param_df2)){
-  param_df2[i,2] <- lambda_predict2(m=lambda_model2(),
-                                  snow=param_df2[i,1])
-}
-plot_1 <- ggplot(param_df2,aes(x=snow_cover,y=lambda)) +
-          geom_point() +
-          xlab("Snow Cover") +
-          ylab("Predicted Lambda")
-
-plot(plot_1)
-```
-
-These results look much more reasonable.
-The maximum predicted $\lambda$ is only 2.5, which is still high, but not unreasonable.
-And positive population growth ($\lambda > 1$) is predicted for snow cover values from \~ 35% to 99%.
-
-And, for comparison, here is a plot of the original data from which the prediction equation was derived:
-
-```{r}
-df <- data.frame(Snow.Cover=df_clean$Snow.Cover, Lambda=df_clean$r)
-plot_2 <- ggplot(df,aes(x=Snow.Cover,y=Lambda)) +
-          geom_point() +
-          xlab("Snow Cover") +
-          ylab("Observed Lambda")
-
-plot(plot_2)
-```
-
-## Remaining questions and issues
-
-1.
-Obviously, the biggest problem is what to do about these environmental variables like degree days that do not have well-behaved quadratic functions.
-I wonder if this is an artifact caused by snow cover being manipulated, but environmental variables not being controlled.
-However, I thought that Alden had provided me with just data from all the control plots, so I don't know if this is the issue.
-I was really hoping we could have at least two predictor variables in the regression model, but that may not work for this data set.
-
-2.
-In order for Matt to use this model for continental projections, we will need snow cover for each pixel in the continental landscape.
-From talking with Alden, it sounds like those data will be available, although I don't know much about the grain sizes and resolution.
-
-3.
-In the first set of regression models, there is a strong effect of initial density, suggesting density dependence, but both the linear and quadratic coefficients are positive, so this is actually some form of inverse density dependence.
-I can't tell if this is an artifact, or if the density effects are what is actually driving the other patterns that we are seeing.
-But we need to discuss this more.
-
-4.
-Alden's data set has additional columns indicating snow conditions (Upwind, downwind, plus, minus).
-I have ignored those designations in this analysis, but they obviously contribute to variation in $\lambda$.
-If you want, it would be easy enough to go back and subscript the data to fit to fewer points.
-That will reduce the variance, although I don't think it is going to change the shape of the curves that much.
-
-5.
-As you can see, the basic model with snow cover looks reasonable, but we need to figure out what, if anything to do with the other environmental variables.
-Once we get this sorted out and agree on a final regression model, I will clean up the code so that Matt has a single function to work with for generating maps.
-
-6. Michael L. and Alden (and Jen??), we should probably talk again on skype after you have digested these results. I will hold off on any other analyses until we have talked.
+Can anyone figure out what is going on with these data?
+I suppose I could just ignore it because we are going to use the linear density model, but it makes me uneasy to see this kind of parameter fit.
+Your comments on this issue will be most appreciated!

### 2024-11-08 Scripts/HouseSparrowSplinefitter.R

@@ -108,6 +108,24 @@ plot(tooWarm, col=rgb(255/255,134/255,116/255, 0.5), legend=F, add=T)
 dev.off()
 
 
+# house sparrow
+housesparrow <- read.csv("/Users/mfitzpatrick/code/InvasionModels/Graphics/houseSparrow/houseSparrow_subSample.csv")
+
+#png(filename="/Users/mfitzpatrick/code/InvasionModels/Graphics/gbif_houseSparrow_tempMean.png",
+#    width=6, height=6, units="in", res=300)
+tiff(filename="/Users/mfitzpatrick/code/InvasionModels/Graphics/gbif_houseSparrow.tif", width=12, height=12, units="in", res=300, 
+     compression="lzw", type="cairo")
+#plot(pos, main="House sparrow population growth (mean temp)", legend=F, col="gray80")
+plot(tmean_bs, col="gray50", legend=F)
+#plot(tooCold, col="gray50", add=T, legend=F)
+#plot(tooWarm, col="gray50", add=T, legend=F)
+points(housesparrow$decimalLongitude, 
+       housesparrow$decimalLatitude, pch=20, 
+       col=rgb(0,0,1,0.25), cex=0.5)
+dev.off()
+
+
+
 # individual months within the breeding season
 tmean_mar <- mean(tmn[[1]], tmx[[1]])
 tmean_apr <- mean(tmn[[2]], tmx[[2]])

### 2024-11-08 Scripts/runModels_plotPredictions.R

@@ -107,7 +107,7 @@ pred_brom_lam <- function(new=NULL,f=model_reduced2){
   return(z)}
 
 # load & prep data 
-df <- read.csv("/Users/mfitzpatrick/code/RoL/Bromus_Clean_V3_Oct-08-19.csv", stringsAsFactors=FALSE)
+df <- read.csv("/Users/mfitzpatrick/code/InvasionModels/CleanData/Bromus_Clean_V3_Oct-08-19.csv", stringsAsFactors=FALSE)
 vifstep(df[,9:18])
 df <- df[,-ncol(df)]                 # remove empty final column
 df_clean <- df[complete.cases(df),]  # remove 4 missing lambda
@@ -147,8 +147,8 @@ model_reduced2 <- lm(data=final_df,r~poly(snow_cover,2, raw=TRUE) +
 
 
 # Extract raster data, predict model and map output ----------------------------
-snow_cover <- raster("/Volumes/Samsung_T5/Projects/misc/RoL/sweProportion.tif")
-win_precip_rain <- raster("/Volumes/Samsung_T5/Projects/misc/RoL/winterRain.tif")
+snow_cover <- raster("/Volumes/dataSSD/Projects/activeProjects/misc/RoL/sweProportion.tif")
+win_precip_rain <- raster("/Volumes/dataSSD/Projects/activeProjects/misc/RoL/winterRain.tif")
 
 # deal with NA values in rasters
 NAs <- unique(c(which(is.na(snow_cover[])), which(is.na(win_precip_rain[]))))
@@ -170,11 +170,13 @@ predRast[predRast[]<=0] <- NA
 
 # plot bromus prediction -------------------------------------------------------
 # North America, with political borders
-NAstates <- shapefile("/Volumes/Samsung_T5/Projects/plantGenome/NA_wIslands_states.shp")
-NAstates.simp <- gSimplify(NAstates, 0.01)
+#NAstates <- shapefile("/Volumes/dataSSD/Projects/activeProjects/plantGenome/NA_wIslands_states.shp")
+#NAstates.simp <- gSimplify(NAstates, 0.01)
+#shapefile(NAstates.simp, filename="/Volumes/dataSSD/Projects/activeProjects/plantGenome/NA_wIslands_states_SIMPLE.shp")
+NAstates.simp <- shapefile("/Volumes/dataSSD/Projects/activeProjects/plantGenome/NA_wIslands_states_SIMPLE.shp")
 
 # lakes
[…]

### 2024-11-08 Scripts/maxentModeling.R
add terraclimate

@@ -1,12 +1,61 @@
 library(dismo)
-library(mapdata)
+library(rnaturalearth)
+library(terra)
+library(sf)
+library(geodata)
+library(dplyr)
+library(ggplot2)
+library(ncdf4)
+library(QBMS)
 
-# medfly
-medfly <- gbif(genus="Ceratitis", 
-               species="capitata", 
-               geo=T, 
-               removeZeros=T)#,
-#ext=extent(c(-180, -30, 0, 90)))
+
+# medfly -----------------------------------------------------------------------
+# download worldclim monthly temperature data
+tavg <- worldclim_global(var='tavg', res=2.5, path="/home/mfitzpatrick/Projects/InvasionModels/")
+
+# medfly occurrence data from GBIF
+medfly <- sp_occurrence(genus='Ceratitis', species='capitata',
+                          download=TRUE, geom=TRUE, 
+                          removeZeros = TRUE) %>%
+  # select columns of interest
+  select(species, country, lon, lat, year, month, coordinateUncertaintyInMeters) %>%
+  # remove high uncertainty records
+  filter(coordinateUncertaintyInMeters < 5000) %>% # 2.5 arc-min = ~ 5km
+  # remove records before 2000
+  #filter(year >= 1970 & year <= 2000) %>%
+  # remove duplicates
+  distinct(lon, lat, .keep_all = TRUE) %>%
+  # convert to sf object
+  st_as_sf(coords = c("lon", "lat"), crs = 4326)
+
+# ggplot of records 
+ggplot() +
+  geom_sf(data = rnaturalearth::ne_countries(scale = "medium", returnclass = "sf")) +
+  geom_sf(data = medfly, aes(color = year)) +
+  theme_minimal()
+
+# function to extract temperature data from terraclimate
+
+get_terraclimate(
+  -1.59,
+  -77.71,
+  from = "1958-01-01",
+  to = "1960-01-01",
+  clim_vars = "tmax",
+  month_mask = 1,
+  offline = FALSE)
+
+
+# read in the full period of record using aggregated files
+
+data <- as.numeric(ncvar_get(nc, varid = var, start = start, count))
+
+
+#remove spatial duplicates:  
+mf_cells <- cellFromXY(tavg, st_coordinates(medfly)) #identify cells from raster for all occurrences
+mf_dup <- duplicated(mf_cells) #which cells have multiple occurrences (ie spatial duplicates)
[…]
